---
title: "MECOM KO analysis"
author: "Marina Nikolova"
date: "17/05/2023"
output: html_document
---

Interesting articles to consider:
- report of different effects of MECOM (EVI1 on cell cycle depending on the cell type: Overexpression of EVI1 interferes with cytokinesis and leads to accumulation of cells with supernumerary centrosomes in G0/1 phase. Karakaya et al. 2012 https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3467026/ 
- good review; also highlights the controversial findings in MECOM overexpression and KO/KD studies: https://www.sciencedirect.com/science/article/pii/S0378111907001849#fig2

From the data:
EVI1 suppresses CEBPE. CEBPE suppresses MYC. Therefore, MECOM (EVI1) KO is expected to lead to MYC upregualtion and this is what we see
Ref: Glass et al. 2013. Global Identification of EVI1 Target Genes in Acute Myeloid Leukemia. PLoS one


# Setup
```{r setup, include=FALSE, echo = FALSE}

# set working directory for the whole Rmarkdown (with setwed () it would be just for the current chunk)
knitr::opts_knit$set(root.dir = "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/")

```

Load packages.
```{r load packages, echo = FALSE}

library(Seurat)
library(tidyverse)
library(RCurl)
library(cowplot)
library(patchwork)
library(dplyr)
library(ggplot2)
library(colorspace)
library(SeuratWrappers)
library(presto)
library(ggrepel)
library(simspec)
library(ggrepel)
library(uwot)
library(clusterProfiler)
library(enrichplot)
library("org.Hs.eg.db")
library("AnnotationDbi")
library(ReactomePA)
library(forcats)

source("/home/marinani/Scripts/Data_analysis/feature_plots.r")
source("/home/marinani/Scripts/Data_analysis/r_util.r")
source("/home/marinani/Scripts/Data_analysis/differential_expression.r")
source("/links/groups/treutlein/USERS/zhisong_he/Tools/scripts/perturb_prob.r")

```


Load data
```{r load seurat object}

seurat <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/MECOM_KO_css_integrated_seurat.rds")
ec <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/MECOM_KO_css_integrated_seurat_endothelial.rds")
meso <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/MECOM_KO_css_integrated_seurat_mesoderm.rds")
mural <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/MECOM_KO_css_integrated_seurat_mural.rds")

write.csv(VariableFeatures(seurat), "Plots/variable_features.csv")

```

```{r generate needed color palettes and levels}

show_col(c(c("#1F618D","#2874A6"),
                     colorRampPalette(c("#138D75","#45B39D","#73C6B6"))(5),
                     colorRampPalette(c("#F7DC6F","#F1C40F","#B7950B"))(4),
                     colorRampPalette(c("#F5B7B1","#E74C3C","#B03A2E"))(7), 
                     c("#BFC9CA","#566573","#273746")))

# Define colors for samples
mecom_ko_levels <- factor(c("wt_1_day7", "wt_2_day7", #"A2_day7", 
                            "B1_17_1_day7", "B1_17_2_day7", "D2_22_1_day7", "D2_22_2_day7", #"G8_1_1_day7", 
                            "wt_1_day21", "wt_2_day21", "wt_3_day21", "wt_4_day21", "B1_17_1_day21", "B1_17_2_day21", "D2_22_1_day21", "D2_22_2_day21", "D2_22_3_day21" #, "G8_1_1_day21"
                            ))
col_sample <-setNames(c(c("#F39229", "#F07F27", #"#ED6D26", 
                          "#E9C92E", "#D5C62A", "#42B8AE", "#3EB3BD", #"#7378B7", 
                          "#EA5D27", "#E5512C", "#E04432", "#DC3838", "#CBC528", "#A8C130", "#3EADC8", "#3EA7D2", "#489CD3" #, "#8966A9"
                          )), mecom_ko_levels)

# Define colors for RNA_css_snn_res.0.4
avg_av_cond_cl <- sapply(levels(seurat$RNA_css_snn_res.0.4), function(x) rowMeans(seurat@assays$RNA@data[,which(seurat$RNA_css_snn_res.0.4==x)]))
hcl_cl <- hclust(as.dist(1-qlcMatrix::corSparse(avg_av_cond_cl[VariableFeatures(seurat),])), method="ward.D2")
cols_cl <- setNames(colorRampPalette(c("#8966A9", "#0A4A3D", "#876E08", 
                                       "#F7DC6F", "#F1C40F", "#B7950B",  
                                       "#E74C3C","#B03A2E",
                                       "#73C6B6", "#43A995","#138D75", "#0E6B59"))(length(levels(seurat$RNA_css_snn_res.0.4))), levels(seurat$RNA_css_snn_res.0.4)[hcl_cl$order])
res.0.4_levels <- c(11,10,2,3,0,1,7,8,9,6,4,5)
cl_levels <- c(9,6,5,4,10,0,1,3,2,11,7,8)

# Define colors for RNA_css_snn_res.0.1
avg_av_cond_cl <- sapply(levels(seurat$RNA_css_snn_res.0.1), function(x) rowMeans(seurat@assays$RNA@data[,which(seurat$RNA_css_snn_res.0.1==x)]))
hcl_cl <- hclust(as.dist(1-qlcMatrix::corSparse(avg_av_cond_cl[VariableFeatures(seurat),])), method="ward.D2")
cols_cl <- setNames(colorRampPalette(c("#8966A9","#E74C3C","#B03A2E","#F7DC6F","#F5D44F", "#F3CC2F", "#F1C40F", "#DDB40D", "#CAA40C", "#B7950B","#138D75", "#279C85", "#3BAB95", "#4EB6A2", "#60BEAC", "#73C6B6"))(length(levels(seurat$RNA_css_snn_res.0.1))), levels(seurat$RNA_css_snn_res.0.1)[hcl_cl$order])
res.0.1_levels <- c(15,6,10,3,8,0,1,13,2,9,14,11,12,7,4,5)

```


# Find cluster markers

```{r find cluster markers res.0.4}

Idents(seurat) <- "RNA_css_snn_res.0.4"

table(seurat$orig.ident, seurat$RNA_css_snn_res.0.4)

cl_markers_res.0.4 <- FindAllMarkers(seurat, min.pct = 0.25, logfc.threshold = log(1.2))
  
top2_cl_markers_res.0.4 <- cl_markers_res.0.4 %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
top3_cl_markers_res.0.4 <- cl_markers_res.0.4 %>% group_by(cluster) %>%  slice_max(n = 3, order_by = avg_log2FC)
top10_cl_markers_res.0.4 <- cl_markers_res.0.4 %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.4 <- cl_markers_res.0.4 %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.4 <- cl_markers_res.0.4 %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.4, "Plots/top_cl_markers_res.0.4.csv")
write.csv(top50_cl_markers_res.0.4, "Plots/top50_cl_markers_res.0.4.csv")

top_cl_markers_unique <- top3_cl_markers_res.0.4[!duplicated(top3_cl_markers_res.0.4[ , "gene"]), ]  # Delete rows
top2_cl_markers_unique <- top_cl_markers_unique %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

# Remove mitochondria and ribosome genes from the list
blacklist <- c(grep("^MT-", cl_markers_res.0.4$gene, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(cl_markers_res.0.4$gene, blacklist)
cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4 %>% filter(gene %in% diff_genes)
nrow(cl_markers_res.0.4_no_MT_RP)

# Redo top marker selection
top2_cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC) 
top5_cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 5, order_by = avg_log2FC) 
top10_cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.4_no_MT_RP <- cl_markers_res.0.4_no_MT_RP %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.4_no_MT_RP, "Plots/top_cl_markers_res.0.4_no_MT_RP.csv")
write.csv(top50_cl_markers_res.0.4_no_MT_RP, "Plots/top50_cl_markers_res.0.4_no_MT_RP.csv")

```

#### Plot cluster markers

```{r heatmap top markers by cluster res.0.4, fig.height=7, fig.width=4}

Idents(seurat) <- "RNA_css_snn_res.0.4"
levels(seurat) <- cl_levels
seurat$RNA_css_snn_res.0.4 <- factor(seurat$RNA_css_snn_res.0.4, levels=cl_levels)

top2_cl_markers_unique <- top2_cl_markers_unique %>% group_by(cluster)  %>% arrange(factor(cluster, levels=cl_levels))

library(gplots)
avg_cond_cl <- sapply(levels(seurat$RNA_css_snn_res.0.4), function(x) rowMeans(seurat@assays$RNA@data[,which(seurat$RNA_css_snn_res.0.4==x)]))
# scale the data as 0 to 1 (instead)
# for top 2 markers
mat_norm <- apply(avg_cond_cl[top2_cl_markers_unique$gene,], 1, function(x) (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)))
col.order <- top2_cl_markers_unique$gene
mat_norm <- mat_norm[,col.order]

pdf("Plots/heatmap_scaled_top2_markers_cl_res.0.4.pdf", width = 5, height = 14)
heatmap.2(t(mat_norm), Colv=NA, Rowv = NA, trace="none", scale="none", margins=c(7,9), cexRow=1, cexCol=1, col =  colorRampPalette(c("white", "black"))(91), key=T, keysize = 2, density.info = "none") 
dev.off()

```


```{r cell proportions within all cells}

cl_levels <- c(9,6,5,4,10,3,0,1,2,11,7,8)

png("Plots/cell_prop_RNA_css_snn_res.0.4_orig.ident.png", width = 2450, height = 2250, res=500)
df <- as.data.frame(table(seurat$RNA_css_snn_res.0.4, seurat$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.4"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.4, level=cl_levels), y=Freq, x=orig.ident)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of cells") +
    scale_fill_manual(values= cols_cl) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


png("Plots/cell_prop_RNA_css_snn_res.0.4_sample_timepoint.png", width = 2450, height = 2250, res=500)
df <- as.data.frame(table(seurat$RNA_css_snn_res.0.4, seurat$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.4"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.4, level=cl_levels), y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of cells") +
    scale_fill_manual(values= cols_cl) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


df <- as.data.frame(table(seurat$RNA_css_snn_res.0.4, seurat$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.4"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

png("Plots/cell_prop_orig.ident_RNA_css_snn_res.0.4.png", width = 2450, height = 2250, res=500)
ggplot(df, aes(fill=orig.ident, y=Freq, x=RNA_css_snn_res.0.4)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of cells") +
    scale_fill_manual(values= col_sample) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


df <- as.data.frame(table(seurat$Phase, seurat$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "Phase"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

png("Plots/cell_prop_sample_timepoint_cell_phase.png", width = 1550, height = 1850, res=500)
ggplot(df, aes(fill=Phase, y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of cells") +
    scale_fill_manual(values= c("#bdbdbd", "#737373",  "#252525")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()

```


# Perform label transfer from the time course data

```{r label transfer from time course}

# Load the time course seurat
ref <- readRDS("/links/groups/treutlein/USERS/zhisong_he/Work/vascular_organoids/data/data.seurat_untransplanted_noWTC.rds")
levels(ref$annot_celltype) <- as.factor( c(0,3,4,9,10,13,2,7,1,8,16,14,15,18,11,5,6,20,17,19,12))

# Prep the color palette
ref_col <- setNames(c(c("#1F618D","#2874A6"),                                # PSC colors
                     colorRampPalette(c("#138D75","#45B39D","#73C6B6"))(5), # mesoderm colors
                     colorRampPalette(c("#F7DC6F","#F1C40F","#B7950B"))(4), # mural colors
                     colorRampPalette(c("#F5B7B1","#E74C3C","#B03A2E"))(7), # endothelial colors
                     c("#BFC9CA","#566573","#273746")),                     # other cluster colors
                   c(0,3,                # PSC clusters
                     4,9,10,13,2,        # mesoderm clusters
                     7,1,8,16,           # mural cell clusters
                     14,15,18,11,5,6,20, # endothelial cell clusters
                     17,19,12))          # other clusters

# Plot reference data 
plot1 <- DimPlot(ref, group.by="day", reduction="umap_css")  & NoAxes()
plot2 <- DimPlot(ref, group.by="annot_celltype", reduction="umap_css") & NoAxes()
plot3 <- FeaturePlot(ref, c("PDGFRB","PECAM1","DCN","KDR","TFPI2","MKI67"),reduction = "umap_css", order = T, ncol=3, pt.size = 0.1, cols = blue_colscheme(10)) & NoAxes() & NoLegend()
((plot1 / plot2) | plot3) + plot_layout(width = c(1,3))
DimPlot(ref, group.by="RNA_css_snn_res.0.8", reduction="umap_css", cols = ref_col)

# Perform Seurat-based label transfer
anchors <- FindTransferAnchors(reference = ref, query = seurat, dims = 1:30, npcs = 30, reduction = "rpca", reference.reduction = "pca")
predictions <- TransferData(anchorset = anchors, refdata = ref$RNA_css_snn_res.0.8, dims = 1:30)
seurat$celltype_transfer <- predictions$predicted.id
levels(seurat$celltype_transfer) <- as.factor( c(0,3,4,9,10,13,2,7,1,8,16,14,15,18,11,5,6,20,17,19,12))

# Plot Seurat label transfer
plot1 <- DimPlot(seurat, reduction = "umap_css", group.by = "orig.ident", label = F, label.size = 4, repel = T, pt.size = 1.3, cols=col_sample) & NoAxes() 
plot2 <- DimPlot(seurat, reduction = "umap_css", group.by="celltype_transfer", label=T,  cols = ref_col, pt.size = 1.3)
plot3 <- DimPlot(ref, group.by="RNA_css_snn_res.0.8", reduction="umap_css", label=T,  cols = ref_col, pt.size = 1.3)
plot4 <- DimPlot(seurat, reduction = "umap_css", group.by = "RNA_css_snn_res.0.4",cols = col_ct, label = F, label.size = 4, repel = T, pt.size = 1.3) & NoAxes()
plot1 | plot2
plot2 | plot3
plot2 | plot4

# Feature plots of some marker genes
FeaturePlot(seurat, c("PDGFRB","PECAM1","DCN","MECOM","TFPI2","ETV2"),reduction = "umap_css", order = T, ncol=3, pt.size = 0.1, cols = blue_colscheme(10)) & NoAxes() & NoLegend()
FeaturePlot(ref, c("PDGFRB","PECAM1","DCN","MECOM","TFPI2","ETV2"),reduction = "umap_css", order = T, ncol=3, pt.size = 0.1, cols = blue_colscheme(10)) & NoAxes() & NoLegend()


# Project the query data onto the reference UMAP structure
seurat <- MapQuery(anchorset = anchors, reference = ref, query = seurat,
    refdata = list(celltype = "RNA_css_snn_res.0.8"), reference.reduction = "pca", reduction.model = "umap_css")


plot5 <- DimPlot(seurat, reduction = "ref.umap", group.by = "predicted.celltype", label = TRUE,
    label.size = 3, cols = ref_col,repel = TRUE) + NoLegend() + ggtitle("Query transferred labels by transfered cluster")
plot6 <- DimPlot(seurat, reduction = "ref.umap", group.by = "orig.ident", label = TRUE,
    label.size = 3, cols = col_sample,repel = TRUE) + NoLegend() + ggtitle("Query transferred labels by sample")
plot7 <- DimPlot(seurat, reduction = "ref.umap", group.by = "orig.ident", split.by = "orig.ident", label = TRUE,
    label.size = 3, cols = col_sample,repel = TRUE) + NoLegend() + ggtitle("Query transferred labels by sample")
plot3 + plot5 + plot6
plot7
DimPlot(seurat, reduction = "ref.umap", group.by = "RNA_css_snn_res.0.4", label = TRUE,pt.size=1, label.size = 3, cols = col_ct,repel = TRUE, split.by = "RNA_css_snn_res.0.4") + NoLegend() + ggtitle("Query transferred labels by sample")
DimPlot(seurat, reduction = "ref.umap", group.by = "RNA_css_snn_res.0.4", label = TRUE,pt.size=1, label.size = 3, cols = col_ct,repel = TRUE) + NoLegend() + ggtitle("Query transferred labels by sample")

DimPlot(ref, reduction = "umap_css", group.by = "timepoint", split.by = "timepoint", label = F, label.size = 4, repel = T) & NoAxes()

pdf("Plots/umap_label_transfer_timecourse.pdf", width = 11, height = 6)
plot2 | plot3
dev.off()


# create a dataset
df <- as.data.frame(table(seurat$celltype_transfer, seurat$orig.ident))
df$Var2 <- factor(x = df$Var2, levels = mecom_ko_levels)
colnames(df)[which(names(df) == "Var1")] <- "celltype_transfer"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

pdf("Plots/stckd_barplot_cluster_per_sample_celltype_transfer.pdf", width = 5, height = 6)
ggplot(df, aes(fill=factor(celltype_transfer, levels = c(
                     0,3,                # PSC clusters
                     4,9,10,13,2,        # mesoderm clusters
                     7,1,8,16,           # mural cell clusters
                     14,15,18,11,5,6,20, # endothelial cell clusters
                     17,19,12)), y=Freq, x=orig.ident)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proporion of cells") +
    scale_fill_manual(values= ref_col) +
   theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16)) + 
    labs(fill='Cluster based on label transfer') 
dev.off()


# create a dataset
df <- as.data.frame(table(seurat$celltype_transfer, seurat$condition))
df$Var2 <- factor(x = df$Var2, levels = c("KO","WT"))
colnames(df)[which(names(df) == "Var1")] <- "celltype_transfer"
colnames(df)[which(names(df) == "Var2")] <- "condition"

pdf("Plots/stckd_barplot_cluster_per_condition_celltype_transfer.pdf", width = 5, height = 6)
ggplot(df, aes(fill=factor(celltype_transfer, levels = c(
                     0,3,                # PSC clusters
                     4,9,10,13,2,        # mesoderm clusters
                     7,1,8,16,           # mural cell clusters
                     14,15,18,11,5,6,20, # endothelial cell clusters
                     17,19,12)), y=Freq, x=condition)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Condition") + 
    ylab("Proporion of cells") +
    scale_fill_manual(values= ref_col) +
   theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16)) + 
    labs(fill='Cluster based on label transfer') 
dev.off()


```

# Differential expression analysis between conditions

#### Basic DE analysis
```{r general similarity}

# General transcriptomic similarity between clone BVOs
pdf("Plots/hclust_seurat_integrated.pdf", width = 5, height = 8)
Idents(seurat) <- "orig.ident"

avg_expr_sample <- sapply(sort(unique(seurat$orig.ident)), function(orig.ident)
rowMeans(seurat@assays$RNA@data[,which(seurat$orig.ident == orig.ident)]))
colnames(avg_expr_sample) <- mecom_ko_levels
corr_sample <- cor(avg_expr_sample[VariableFeatures(seurat),])
corr_sample
plot(hclust(as.dist(1-corr_sample)))
table(seurat$orig.ident)
dev.off()


pdf("Plots/hclust_ECs.pdf", width = 5, height = 8)
# ECs only
Idents(ec) <- "orig.ident"

avg_expr_sample <- sapply(sort(unique(ec$orig.ident)), function(orig.ident)
rowMeans(ec@assays$RNA@data[,which(ec$orig.ident == orig.ident)]))
colnames(avg_expr_sample) <- mecom_ko_levels
corr_sample <- cor(avg_expr_sample[VariableFeatures(ec),])
corr_sample
plot(hclust(as.dist(1-corr_sample)))
table(ec$orig.ident)
dev.off()

# mural cells only
pdf("Plots/hclust_mural_cells.pdf", width = 5, height = 8)
Idents(mural) <- "orig.ident"

avg_expr_sample <- sapply(sort(unique(mural$orig.ident)), function(orig.ident)
rowMeans(mural@assays$RNA@data[,which(mural$orig.ident == orig.ident)]))
colnames(avg_expr_sample) <- mecom_ko_levels
corr_sample <- cor(avg_expr_sample[VariableFeatures(mural),])
corr_sample
plot(hclust(as.dist(1-corr_sample)))
table(mural$orig.ident)
dev.off()

```

#### Perform wlicoxon test
```{r wilcoxon test on all cells}

Idents(seurat) <- "condition"
# Run wilcoxon test
DE_wilcoxauc_seurat <- wilcoxauc(seurat, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_seurat <- DE_wilcoxauc_seurat %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_seurat) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_seurat$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(DE_wilcoxauc_seurat$feature,blacklist)
DE_wilcoxauc_seurat <- DE_wilcoxauc_seurat %>% filter(feature %in% diff_genes)
nrow(DE_wilcoxauc_seurat)

# plot
pdf("Plots/volcano_DE_all.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_seurat, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(shape=21, size=2, aes(fill=factor(outcome)), position = "jitter") +
  geom_text_repel(max.overlaps = 50) +
  geom_vline(xintercept=c(-log(1.1), log(1.1), 0),  linetype="dotted") +
  geom_hline(yintercept=-log10(0.01), linetype="dotted") +
  theme_classic() + 
  scale_fill_manual(values = c("#F1C619", "gray50", "#D44637")) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_seurat, "Plots/DE_wilcoxauc_seurat.csv")

```

```{r wilcoxon test all clusters separately RNA_css_snn_res.0.4}

Idents(seurat) <- "RNA_css_snn_res.0.4"
# Split seurat object into a list based on the cluster in RNA_css_snn_res.0.1
seurat.clusters <- SplitObject(seurat, split.by = "RNA_css_snn_res.0.4")

# Generate a list of data frames with DEGs
seurat_de_tab <- lapply(seurat.clusters, function(cl){
  seurat_de_tab <- wilcoxauc(cl, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
  blacklist <- c(grep("^MT-", seurat_de_tab$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
  diff_genes <- setdiff(seurat_de_tab$feature,blacklist)
  seurat_de_tab <- seurat_de_tab %>% filter(feature %in% diff_genes)
})

seurat_de_tab <- seurat_de_tab[c("9", "6", "5", "4", "10", "0",  "1",  "3",  "2",  "11", "7",  "8" )]

# Add number of DEGs 
seurat_no.up <- c()
seurat_no.down <- c()
for (i in seurat_de_tab) {
  seurat_no.up <- rbind(seurat_no.up, sum(i$outcome == "Up-regulated"))
  seurat_no.down <- rbind(seurat_no.down, sum(i$outcome ==  "Down-regulated"))
}

# Createa a table with no. DEGs
seurat_no.DEGs <- matrix(c("9", "6", "5", "4", "10", "0",  "1",  "3",  "2",  "11", "7",  "8"), ncol=1, byrow=TRUE)
seurat_no.DEGs <- as.data.frame(seurat_no.DEGs)
colnames(seurat_no.DEGs) <- c("cluster")

seurat_no.DEGs$no.up <- seurat_no.up
seurat_no.DEGs$no.down <- seurat_no.down
seurat_no.DEGs$no.total <- seurat_no.down + seurat_no.up
# seurat$RNA_css_snn_res.0.4 <- factor(seurat$RNA_css_snn_res.0.4, levels=c(0,1,2,3)) # Change the order just for ease addition to the table
seurat_no.DEGs$no.cells <- as.integer(table(seurat$RNA_css_snn_res.0.4))
# seurat$RNA_css_snn_res.0.4 <- factor(seurat$RNA_css_snn_res.0.4, levels=c(3,2,0,1))  # Return to the original order for future use
seurat_no.DEGs <- seurat_no.DEGs[order(-seurat_no.DEGs$no.total),] 
seurat_no.DEGs$cluster <- factor(seurat_no.DEGs$cluster,levels =seurat_no.DEGs$cluster)
#7  5  8  2  1  4  6  3  0  9  10 11

# Define colors
col_order_degs_seurat_cl <- c("#E74C3C","#138D75","#B03A2E","#876F08","#B7950B","#0E6B59","#43A995","#F7DC6F","#F1C40F","#73C6B6","#0A4A3D","#8966A9")

# Plot 
pdf("Plots/seurat_cl_lollipop_DEGs_up.pdf", height = 3, width = 5)
ggplot(seurat_no.DEGs, aes(x = cluster, y = no.up)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.up), col = col_order_degs_seurat_cl) +
  geom_point(aes(x=cluster, y=no.up), size = 8, col = col_order_degs_seurat_cl) +
  geom_text(aes(x=cluster,  y=no.up, label = no.up), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of up-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/seurat_cl_lollipop_DEGs_down.pdf", height = 3, width = 5)
ggplot(seurat_no.DEGs, aes(x = cluster, y = no.down)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.down), col = col_order_degs_seurat_cl) +
  geom_point(aes(x=cluster, y=no.down), size = 8, col = col_order_degs_seurat_cl) +
  geom_text(aes(x=cluster,  y=no.down, label = no.down), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of down-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/seurat_cl_lollipop_DEGs_total.pdf", height = 2, width = 4)
ggplot(seurat_no.DEGs, aes(x = cluster), y = no.total) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.total), col = "gray50") +
  geom_point(aes(x=cluster, y=no.total, size = no.cells), col = col_order_degs_seurat_cl) +
  #geom_text(aes(x=cluster,  y=no.total, label = no.total), color = "black", size = 3) +
  theme_classic() +
  ylim(0,1200) +
  xlab("Cluster") + 
  ylab("Number of DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

write.csv(seurat_de_tab, "Plots/seurat_DE_wilcoxauc_all_cl_separately.csv") # Cluster order is "9", "6", "5", "4", "10", "0",  "1",  "3",  "2",  "11", "7",  "8"
# sapply(names(de_tab), 
#  function (x) write.csv(de_tab[[x]], file=paste("Plots/EC_DE_wilcoxauc_",x, ".csv")))

```

```{r wicoxon test on cluster 8}

Idents(seurat) <- "RNA_css_snn_res.0.4"
cl8 <- subset(seurat, idents = c(8))

# Run wilcoxon test
DE_wilcoxauc_cl8 <- wilcoxauc(cl8, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_cl8 <- DE_wilcoxauc_cl8 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_cl8) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_cl8$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes_cl8 <- setdiff(DE_wilcoxauc_cl8$feature,blacklist)
DE_wilcoxauc_cl8 <- DE_wilcoxauc_cl8 %>% filter(feature %in% diff_genes_cl8)

# plot
pdf("Plots/cl8_volcano_DE.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl8, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/cl8_volcano_DE_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl8, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

# Plot with selected labels
cl8_deg <- c("IGFBP5","NBDY","CRHBP","ITGA1","IL7R","NRP2","FLT4","MAF", "PRCP","CAVIN2","LYVE1","TGFBR2", "ROBO1", "APLNR", "SQLE","UGT3A2","FOS", "NR2F2", "NEFH", "TEK","TGFBR3", # UP in KO
             "CCL2", "CARTPT", "GJA5","GJA4","CAV1","CLDN5","PECAM1","COL1A2","KLF2","IGFBP2","HES4","CXCR4","TAGLN2","DDIT4","EMP1" # UP in WT
             )
# Filter them from the data frame
cl8_deg <- DE_wilcoxauc_cl8 %>% filter(feature %in% cl8_deg)


pdf("Plots/cl8_volcano_DE_seleceted_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl8, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_point(data =  cl8_deg, size = 2,shape = 21,fill = "#B03A2E",colour = "#B03A2E")+
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj")) +
  geom_text_repel(data = cl8_deg, mapping = aes(logFC, -log10(padj), label = feature), size = 3, max.overlaps = 50)
dev.off()


png("Plots/cl8_volcano_DE_seleceted_label.png", height = 2000, width = 3000, res = 500)
ggplot(DE_wilcoxauc_cl8, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 0.7) +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Up-regulated" & logFC > log(1.1)), size = 1,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl8, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1,shape = 21,fill = "black",colour = "black")+
  geom_point(data =  cl8_deg, size = 1.2,shape = 21,fill = "#B03A2E",colour = "#B03A2E")+
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj")) 
dev.off()

write.csv(DE_wilcoxauc_cl8, "Plots/DE_wilcoxauc_cl8.csv")

```


```{r wicoxon test on cluster 7}

Idents(seurat) <- "RNA_css_snn_res.0.4"
cl7 <- subset(seurat, idents = c(7))

# Run wilcoxon test
DE_wilcoxauc_cl7 <- wilcoxauc(cl7, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_cl7 <- DE_wilcoxauc_cl7 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_cl7$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes_cl7 <- setdiff(DE_wilcoxauc_cl7$feature,blacklist)
DE_wilcoxauc_cl7 <- DE_wilcoxauc_cl7 %>% filter(feature %in% diff_genes_cl7)

# plot
pdf("Plots/cl7_volcano_DE.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl7, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl7, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl7, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_cl7, "Plots/DE_wilcoxauc_cl7.csv")

```

```{r wicoxon test on cluster 2}

Idents(seurat) <- "RNA_css_snn_res.0.4"
cl2 <- subset(seurat, idents = c(2))

# Run wilcoxon test
DE_wilcoxauc_cl2 <- wilcoxauc(cl2, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_cl2 <- DE_wilcoxauc_cl2 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_cl2$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes_cl2 <- setdiff(DE_wilcoxauc_cl2$feature,blacklist)
DE_wilcoxauc_cl2 <- DE_wilcoxauc_cl2 %>% filter(feature %in% diff_genes_cl2)
DE_wilcoxauc_cl2$padj[DE_wilcoxauc_cl2$padj == 0.000000e+00] <- min(DE_wilcoxauc_cl2$padj[which(DE_wilcoxauc_cl2$padj != 0)])

# plot
pdf("Plots/cl2_volcano_DE_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/cl2_volcano_DE_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  # geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()


# Plot with selected labels
cl2_deg <- c("IGFBP5", "FOS", "NR2F1", "CYP1B2", "NBDY", "CXCL14", "SOX4", "MEST", "TGFBI", "TGFBR2", # UP in KO
             "CRHBP", "TAC1", "DLK1", "IGFBP7","ACTA2","PDGFRB", "COL18A1","HES4","TAGLN","SSL2","MYL9" # UP in WT
             )
# Filter them from the data frame
cl2_deg <- DE_wilcoxauc_cl2 %>% filter(feature %in% cl2_deg)


pdf("Plots/cl2_volcano_DE_seleceted_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_point(data =  cl2_deg, size = 2,shape = 21,fill = "#B03A2E",colour = "#B03A2E")+
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj")) +
  geom_text_repel(data = cl2_deg, mapping = aes(logFC, -log10(padj), label = feature), size = 3, max.overlaps = 50)
dev.off()


png("Plots/cl2_volcano_DE_seleceted_label.png", height = 2000, width = 3000, res = 500)
ggplot(DE_wilcoxauc_cl2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 0.7) +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1,shape = 21,fill = "black",colour = "black")+
  geom_point(data =  cl2_deg, size = 1.2,shape = 21,fill = "#B03A2E",colour = "#B03A2E")+
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj")) 
dev.off()

write.csv(DE_wilcoxauc_cl2, "Plots/DE_wilcoxauc_cl2.csv")

```

```{r wicoxon test on cluster 5}

Idents(seurat) <- "RNA_css_snn_res.0.4"
cl5 <- subset(seurat, idents = c(5))

# Run wilcoxon test
DE_wilcoxauc_cl5 <- wilcoxauc(cl5, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_cl5 <- DE_wilcoxauc_cl5 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_cl5$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes_cl5 <- setdiff(DE_wilcoxauc_cl5$feature,blacklist)
DE_wilcoxauc_cl5 <- DE_wilcoxauc_cl5 %>% filter(feature %in% diff_genes_cl5)

# plot
pdf("Plots/cl5_volcano_DE.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl5, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl5, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl5, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_cl5, "Plots/DE_wilcoxauc_cl5.csv")

```


```{r wicoxon test on clusters 0 1}

Idents(seurat) <- "RNA_css_snn_res.0.4"
cl0_1 <- subset(seurat, idents = c(0,1))

# Run wilcoxon test
DE_wilcoxauc_cl0_1 <- wilcoxauc(cl0_1, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_cl0_1 <- DE_wilcoxauc_cl0_1 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_cl0_1) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_cl0_1$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes_cl0_1 <- setdiff(DE_wilcoxauc_cl0_1$feature,blacklist)
DE_wilcoxauc_cl0_1 <- DE_wilcoxauc_cl0_1 %>% filter(feature %in% diff_genes_cl0_1)

# plot
pdf("Plots/cl0_1_volcano_DE.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl0_1, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl0_1, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl0_1, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/cl0_1_volcano_DE_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_cl0_1, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_cl0_1, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_cl0_1, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_cl0_1, "Plots/DE_wilcoxauc_cl0_1.csv")

```

```{r KEGG of clusters 8 2 5 7 0_1}

# Cluster 8
DE_wilcoxauc_cl8_pos <- DE_wilcoxauc_cl8 %>% filter(outcome != "Unchanged")
DE_wilcoxauc_cl8_pos$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=DE_wilcoxauc_cl8_pos$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")

fm_path_enrich_cl8 <- compareCluster(entrez_id~outcome, data=DE_wilcoxauc_cl8_pos, fun = enrichKEGG, organism="hsa")
pdf("Plots/KEGG_enrichment_res_cl8.pdf", width = 4, height = 5)
enrichplot::dotplot(fm_path_enrich_cl8, showCategory = 6) +
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(fm_path_enrich_cl8@compareClusterResult, "Plots/KEGG_path_enrich_cl8.csv")


# Cluster 2
DE_wilcoxauc_cl2_pos <- DE_wilcoxauc_cl2 %>% filter(outcome != "Unchanged")
DE_wilcoxauc_cl2_pos$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=DE_wilcoxauc_cl2_pos$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")

fm_path_enrich_cl2 <- compareCluster(entrez_id~outcome, data=DE_wilcoxauc_cl2_pos, fun = enrichKEGG, organism="hsa")
pdf("Plots/KEGG_enrichment_res_cl2.pdf", width = 4, height = 5)
enrichplot::dotplot(fm_path_enrich_cl2, showCategory = 6) +
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(fm_path_enrich_cl2@compareClusterResult, "Plots/KEGG_path_enrich_cl2.csv")


# Cluster 7
DE_wilcoxauc_cl7_pos <- DE_wilcoxauc_cl7 %>% filter(outcome != "Unchanged")
DE_wilcoxauc_cl7_pos$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=DE_wilcoxauc_cl7_pos$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")

fm_path_enrich_cl7 <- compareCluster(entrez_id~outcome, data=DE_wilcoxauc_cl7_pos, fun = enrichKEGG, organism="hsa")
pdf("Plots/KEGG_enrichment_res_cl7.pdf", width = 4, height = 5)
enrichplot::dotplot(fm_path_enrich_cl7, showCategory = 6) +
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(fm_path_enrich_cl7@compareClusterResult, "Plots/KEGG_path_enrich_cl7.csv")


# Cluster 5
DE_wilcoxauc_cl5_pos <- DE_wilcoxauc_cl5 %>% filter(outcome != "Unchanged")
DE_wilcoxauc_cl5_pos$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=DE_wilcoxauc_cl5_pos$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")

fm_path_enrich_cl5 <- compareCluster(entrez_id~outcome, data=DE_wilcoxauc_cl5_pos, fun = enrichKEGG, organism="hsa")
pdf("Plots/KEGG_enrichment_res_cl5.pdf", width = 4, height = 5)
enrichplot::dotplot(fm_path_enrich_cl5, showCategory = 6) +  
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(fm_path_enrich_cl5@compareClusterResult, "Plots/KEGG_path_enrich_cl5.csv")


# Clusters 0 and 1
DE_wilcoxauc_cl0_1_pos <- DE_wilcoxauc_cl0_1 %>% filter(outcome != "Unchanged")
DE_wilcoxauc_cl0_1_pos$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=DE_wilcoxauc_cl0_1_pos$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")

fm_path_enrich_cl0_1 <- compareCluster(entrez_id~outcome, data=DE_wilcoxauc_cl0_1_pos, fun = enrichKEGG, organism="hsa")
pdf("Plots/KEGG_enrichment_res_cl0_1.pdf", width = 4, height = 5)
enrichplot::dotplot(fm_path_enrich_cl0_1, showCategory = 6) +  
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(fm_path_enrich_cl0_1@compareClusterResult, "Plots/KEGG_path_enrich_cl0_1.csv")

```

```{r KEGG of all clusters}

seurat_de_tab <- lapply(seurat.clusters, function(cl){
  seurat_de_tab <- wilcoxauc(cl, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
  blacklist <- c(grep("^MT-", seurat_de_tab$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
  diff_genes <- setdiff(seurat_de_tab$feature,blacklist)
  seurat_de_tab <- seurat_de_tab %>% filter(feature %in% diff_genes)
})

seurat_de_tab <- seurat_de_tab[c("9", "6", "5", "4", "10", "3", "0",  "1",  "2",  "11", "7",  "8" )]


# # Add a column with the cluster to each data frame
# seurat_de_tab_pos <- lapply(names(seurat_de_tab), function(x) {
#     seurat_de_tab[[x]] %>% 
#         mutate(Cluster = x)
# })

# Add ENTREZID to each dataframe for each gene
seurat_de_tab_entr <- lapply(seurat_de_tab, function (cl){
  seurat_de_tab_entr <- cl %>% filter(outcome != "Unchanged")
  seurat_de_tab_entr$entrez_id <- mapIds(org.Hs.eg.db,
                    keys=seurat_de_tab_entr$feature,
                    column="ENTREZID",
                    keytype="SYMBOL",
                    multiVals="first")
  return(seurat_de_tab_entr)
})

# Extract a list of gene names of uprefgulated genes in each cluster
seurat_de_tab_up <- lapply(seurat_de_tab_entr, function (cl){
  seurat_de_tab_up <- cl %>% filter(outcome == "Up-regulated")
  seurat_de_tab_up <- seurat_de_tab_up$entrez_id
  return(seurat_de_tab_up)
})

ck <- compareCluster(geneCluster = seurat_de_tab_up, fun = enrichKEGG,)
ck <- setReadable(ck, OrgDb = org.Hs.eg.db, keyType="ENTREZID")
head(ck) 

pdf("Plots/KEGG_enrichment_res_all.pdf", width = 8, height = 8)
enrichplot::dotplot(ck, showCategory = 6) +  
  ggtitle('Pathway enrichment') + 
  scale_colour_gradientn(colours = rev(bluewhitered_colscheme(30))) + 
  RotatedAxis() +
  theme_classic()
dev.off()

write.csv(ck@compareClusterResult, "Plots/KEGG_path_enrich_all_clusters.csv")

```

```{r feature plot known markers}

markers <- c("HAND1", "TWIST1", "PITX1", "MECOM", "TFPI2", "COL6A1",
             "PECAM1", "CLDN5", "CDH5", "KDR", "GJA5", "SOX7",
             "DCN","PDGFRB","ACTA2","LUM","DLK1","IGFBP5",
             "NR2F2","APLNR","EPHB4","CXCR4","GJA4","EFNB2")

png("Plots/featureplots_markers.png", width = 9800, height = 6000, res = 500)
plotMultiFeatures(Embeddings(seurat, "umap_css"),
                  seurat@assays$RNA@data[markers,],
                  ncol=6, colorPal=blue_colscheme, cex=0.5, par_cex = 0.8, mar = c(1,1,2,2), cex.main=1, sort_by_value=T)
dev.off()

```

# Endothelial cells only

#### Find EC cluster markers
```{r EC cluster markers res.0.1}

Idents(ec) <- "RNA_css_snn_res.0.1"

table(ec$orig.ident, ec$RNA_css_snn_res.0.1)

cl_markers_res.0.1 <- FindAllMarkers(ec, min.pct = 0.25, logfc.threshold = log(1.2))
  
top2_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
top3_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 3, order_by = avg_log2FC)
top10_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.1 <- cl_markers_res.0.1 %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.1, "Plots/EC_top_cl_markers_res.0.1.csv")
write.csv(top50_cl_markers_res.0.1, "Plots/EC_top50_cl_markers_res.0.1.csv")

# top_cl_markers_unique <- top3_cl_markers_res.0.1[!duplicated(top3_cl_markers_res.0.1[ , "gene"]), ]  # Delete rows
# top2_cl_markers_unique <- top_cl_markers_unique %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

# Remove mitochondria and ribosome genes from the list
blacklist <- c(grep("^MT-", cl_markers_res.0.1$gene, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(cl_markers_res.0.1$gene, blacklist)
cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1 %>% filter(gene %in% diff_genes)
nrow(cl_markers_res.0.4_no_MT_RP)

# Redo top marker selection
top2_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC) 
top5_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 5, order_by = avg_log2FC) 
top10_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.1_no_MT_RP, "Plots/EC_top_cl_markers_res.0.1_no_MT_RP.csv")
write.csv(top50_cl_markers_res.0.1_no_MT_RP, "Plots/EC_top50_cl_markers_res.0.1_no_MT_RP.csv")

```

#### Plot EC cluster markers

```{r heatmap EC top markers by cluster res.0.1}

EC_top_cl_markers_res.0.1_no_MT_RP <- read.csv("Plots/Endothelial_cells/EC_top_cl_markers_res.0.1_no_MT_RP.csv")
EC_top2_cl_markers_res.0.1_no_MT_RP <- EC_top_cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC) 

ec_levels <- c(1,0,2,3)

Idents(ec) <- "RNA_css_snn_res.0.1"
levels(ec) <- ec_levels
ec$RNA_css_snn_res.0.1 <- factor(ec$RNA_css_snn_res.0.1, levels=ec_levels)

EC_top2_cl_markers_res.0.1_no_MT_RP <- EC_top2_cl_markers_res.0.1_no_MT_RP %>% group_by(cluster)  %>% arrange(factor(cluster, levels=ec_levels))

avg_cond_cl <- sapply(levels(ec$RNA_css_snn_res.0.1), function(x) rowMeans(ec@assays$RNA@data[,which(ec$RNA_css_snn_res.0.1==x)]))
# scale the data as 0 to 1 (instead)
# for top 2 markers
mat_norm <- apply(avg_cond_cl[EC_top2_cl_markers_res.0.1_no_MT_RP$gene,], 1, function(x) (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)))
col.order <- EC_top2_cl_markers_res.0.1_no_MT_RP$gene
mat_norm <- mat_norm[,col.order]

pdf("Plots/EC_heatmap_scaled_top2_markers_cl_res.0.1.pdf", width = 4, height = 9)
heatmap.2(t(mat_norm), Colv=NA, Rowv = NA, trace="none", scale="none", margins=c(7,9), cexRow=1, cexCol=1, col =  colorRampPalette(c("white", "black"))(91), key=T, keysize = 2, density.info = "none") 
dev.off()

```

```{r plot EC}

avg_av_cond_cl_ec <- sapply(levels(ec$RNA_css_snn_res.0.1), function(x) rowMeans(ec@assays$RNA@data[,which(ec$RNA_css_snn_res.0.1==x)]))
hcl_cl_ec <- hclust(as.dist(1-qlcMatrix::corSparse(avg_av_cond_cl_ec[VariableFeatures(ec),])), method="ward.D2")
cols_cl_ec <- setNames(colorRampPalette(c( "#EB6F63", "#F5B7B1","#D44637", "#B03A2E"))(length(levels(ec$RNA_css_snn_res.0.1))), levels(ec$RNA_css_snn_res.0.1)[hcl_cl_ec$order])

```

```{r cell proportions within EC clusters}

pdf("Plots/EC_cell_prop_RNA_css_snn_res.0.1_orig.ident.pdf", width = 8, height = 7)
df <- as.data.frame(table(ec$RNA_css_snn_res.0.1, ec$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.1), y=Freq, x=orig.ident)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of endothelial cells") +
    scale_fill_manual(values= cols_cl_ec) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/EC_cell_prop_RNA_css_snn_res.0.1_sample_timepoint.pdf", width = 8, height = 7)
df <- as.data.frame(table(ec$RNA_css_snn_res.0.1, ec$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.1), y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of endothelial cells") +
    scale_fill_manual(values= cols_cl_ec) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/EC_cell_prop_orig.ident_RNA_css_snn_res.0.1.pdf", width = 8, height = 7)
df <- as.data.frame(table(ec$RNA_css_snn_res.0.1, ec$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=orig.ident, y=Freq, x=RNA_css_snn_res.0.1)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of endothelial cells") +
    scale_fill_manual(values= col_sample) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/EC_cell_prop_sample_timepoint_cell_phase.pdf", width = 6, height = 7)
df <- as.data.frame(table(ec$Phase, ec$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "Phase"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=Phase, y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample per timepoint") + 
    ylab("Proportion of endothelial cells") +
    scale_fill_manual(values= c("#bdbdbd", "#737373",  "#252525")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()

```

#### Perform wilcoxon test on all EC clusters RNA_css_snn_res.0.1
```{r wilcoxon test all EC clusters separately RNA_css_snn_res.0.1}

Idents(ec) <- "RNA_css_snn_res.0.1"
# Split ec object into a list based on the cluster in RNA_css_snn_res.0.1
ec.clusters <- SplitObject(ec, split.by = "RNA_css_snn_res.0.1")

# Generate a list of data frames with DEGs
ec_de_tab <- lapply(ec.clusters, function(cl){
  de_tab <- wilcoxauc(cl, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
  blacklist <- c(grep("^MT-", de_tab$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
  diff_genes <- setdiff(de_tab$feature,blacklist)
  de_tab <- de_tab %>% filter(feature %in% diff_genes)
})


# Add number of DEGs 
ec_no.up <- c()
ec_no.down <- c()
for (i in ec_de_tab) {
  ec_no.up <- rbind(ec_no.up, sum(i$outcome == "Up-regulated"))
  ec_no.down <- rbind(ec_no.down, sum(i$outcome ==  "Down-regulated"))
}

# Createa a table with no. DEGs
ec_no.DEGs <- matrix(c("0","1","2","3"), ncol=1, byrow=TRUE)
ec_no.DEGs <- as.data.frame(ec_no.DEGs)
colnames(ec_no.DEGs) <- c("cluster")

ec_no.DEGs$no.up <- ec_no.up
ec_no.DEGs$no.down <- ec_no.down
ec_no.DEGs$no.total <- ec_no.down + ec_no.up
ec$RNA_css_snn_res.0.1 <- factor(ec$RNA_css_snn_res.0.1, levels=c(0,1,2,3)) # Change the order just for ease addition to the table
ec_no.DEGs$no.cells <- as.integer(table(ec$RNA_css_snn_res.0.1))
ec$RNA_css_snn_res.0.1 <- factor(ec$RNA_css_snn_res.0.1, levels=c(3,2,0,1))  # Return to the original order for future use
ec_no.DEGs <- ec_no.DEGs[order(-ec_no.DEGs$no.total),] 
ec_no.DEGs$cluster <- factor(ec_no.DEGs$cluster,levels =ec_no.DEGs$cluster)
#1,0,2,3

# Define colors
col_order_degs_ec <- c("#B03A2E", "#D44637", "#EB6F63", "#F5B7B1") 

# Plot 
pdf("Plots/EC_lollipop_DEGs_up.pdf", height = 7, width = 5)
ggplot(ec_no.DEGs, aes(x = cluster, y = no.up)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.up), col = col_order_degs_ec) +
  geom_point(aes(x=cluster, y=no.up), size = 8, col = col_order_degs_ec) +
  geom_text(aes(x=cluster,  y=no.up, label = no.up), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of up-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/EC_lollipop_DEGs_down.pdf", height = 7, width = 5)
ggplot(ec_no.DEGs, aes(x = cluster, y = no.down)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.down), col = col_order_degs_ec) +
  geom_point(aes(x=cluster, y=no.down), size = 8, col = col_order_degs_ec) +
  geom_text(aes(x=cluster,  y=no.down, label = no.down), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of down-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/EC_lollipop_DEGs_total.pdf", height = 3, width = 4)
ggplot(ec_no.DEGs, aes(x = cluster), y = no.total) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.total), col = "gray50") +
  geom_point(aes(x=cluster, y=no.total, size = no.cells), col = col_order_degs_ec) +
  #geom_text(aes(x=cluster,  y=no.total, label = no.total), color = "black", size = 3) +
  theme_classic() +
  ylim(0,1500) +
  xlab("Cluster") + 
  ylab("Number of DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()


# Volcano plots
pdf("Plots/EC_volcano_DE_KO_vs_WT_all.pdf", height = 5, width = 8) # Cluster order is 0,1,2,3
lapply(ec_de_tab, function(cl) {ggplot(cl, aes(x = logFC, y = -log10(padj),label=DEG)) +
  # geom_point(shape=21, size=2, aes(fill=factor(outcome)), position = "jitter") +
  geom_point(colour = "grey", alpha = 0.5) +
  geom_point(data =  filter(cl, outcome == "Up-regulated" & logFC > log(1.1)), size = 1,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(cl, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
})
dev.off()


write.csv(ec_de_tab, "Plots/EC_DE_wilcoxauc_all.csv") # Cluster order is 0,1,2,3
# sapply(names(de_tab), 
#  function (x) write.csv(de_tab[[x]], file=paste("Plots/EC_DE_wilcoxauc_",x, ".csv")))


## Find total DEGs between ECs
ec_de_all <- wilcoxauc(ec, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
blacklist <- c(grep("^MT-", ec_de_all$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(ec_de_all$feature,blacklist)
ec_de_all <- ec_de_all %>% filter(feature %in% diff_genes)
ec_de_all$padj[ec_de_all$padj == 0.000000e+00] <- min(ec_de_all$padj[which(ec_de_all$padj != 0)])

pdf("Plots/EC_volcano_DE_KO_vs_WT.pdf", height = 5, width = 8)
ggplot(ec_de_all, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(ec_de_all, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(ec_de_all, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

```

```{r wicoxon test on EC clusters 1,0,2 RNA_css_snn_res.0.1}

Idents(ec) <- "RNA_css_snn_res.0.1"
ec_1_0_2 <- subset(ec, idents = c(1,0,2))
Idents(ec_1_0_2) <- "condition"

# Run wilcoxon test
DE_wilcoxauc_ec_1_0_2 <- wilcoxauc(ec_1_0_2, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_ec_1_0_2 <- DE_wilcoxauc_ec_1_0_2 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_ec_1_0_2) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_ec_1_0_2$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(DE_wilcoxauc_ec_1_0_2$feature,blacklist)
DE_wilcoxauc_ec_1_0_2 <- DE_wilcoxauc_ec_1_0_2 %>% filter(feature %in% diff_genes)
DE_wilcoxauc_ec_1_0_2$padj[DE_wilcoxauc_ec_1_0_2$padj == 0.000000e+00] <- min(DE_wilcoxauc_ec_1_0_2$padj[which(DE_wilcoxauc_ec_1_0_2$padj != 0)])

pdf("Plots/EC_volcano_DE_KO_vs_WT_cl1_0_2.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_ec_1_0_2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_ec_1_0_2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_ec_1_0_2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/EC_volcano_DE_KO_vs_WT_cl1_0_2_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_ec_1_0_2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_ec_1_0_2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_ec_1_0_2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  # geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_ec_1_0_2, "Plots/EC_volcano_DE_KO_vs_WT_cl1_0_2.csv") 

```

```{r feature plots EC}

ec_markers <- c("PECAM1","CLDN5","KDR","NR2F2","MECOM","APLNR",
                "EPHB4","GJA4","EFNB2","CXCR4", "ESM1","MYC",
                "TAC1","MEF2C","IGF2.1","CCL2","ICAM2", # based on DE analysis - DOWN in KO
                "CRHBP","IGFBP5","NTS","GJA1","HES4","CD44" # based on DE analysis - UP in KO
                )

png("Plots/EC_featureplots_markers.png", width = 7000, height = 4000, res = 500)
plotMultiFeatures(Embeddings(ec, "umap_css"),
                  ec@assays$RNA@data[ec_markers,],
                  ncol=6, colorPal=blue_colscheme, cex=0.5, par_cex = 0.8, mar = c(1,1,2,2), cex.main=1, sort_by_value=T)
dev.off()

```



# Mesoderm cells only

#### Find mesoderm cluster markers
```{r mesoderm cluster markers res.0.1}

Idents(meso) <- "RNA_css_snn_res.0.1"

table(meso$orig.ident, meso$RNA_css_snn_res.0.1)

cl_markers_res.0.1 <- FindAllMarkers(meso, min.pct = 0.25, logfc.threshold = log(1.2))
  
top2_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
top3_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 3, order_by = avg_log2FC)
top10_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.1 <- cl_markers_res.0.1 %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.1 <- cl_markers_res.0.1 %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.1, "Plots/Meso_top_cl_markers_res.0.1.csv")
write.csv(top50_cl_markers_res.0.1, "Plots/Meso_top50_cl_markers_res.0.1.csv")

# top_cl_markers_unique <- top3_cl_markers_res.0.1[!duplicated(top3_cl_markers_res.0.1[ , "gene"]), ]  # Delete rows
# top2_cl_markers_unique <- top_cl_markers_unique %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

# Remove mitochondria and ribosome genes from the list
blacklist <- c(grep("^MT-", cl_markers_res.0.1$gene, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(cl_markers_res.0.1$gene, blacklist)
cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1 %>% filter(gene %in% diff_genes)
nrow(cl_markers_res.0.4_no_MT_RP)

# Redo top marker selection
top2_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC) 
top5_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 5, order_by = avg_log2FC) 
top10_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.1_no_MT_RP <- cl_markers_res.0.1_no_MT_RP %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.1_no_MT_RP, "Plots/Meso_top_cl_markers_res.0.1_no_MT_RP.csv")
write.csv(top50_cl_markers_res.0.1_no_MT_RP, "Plots/Meso_top50_cl_markers_res.0.1_no_MT_RP.csv")

```

#### Plot mesoderm cluster markers

```{r heatmap mesoderm top markers by cluster res.0.1}

meso_top_cl_markers_res.0.1_no_MT_RP <- read.csv("Plots/Mesoderm_cells/Meso_top_cl_markers_res.0.1_no_MT_RP.csv")
meso_top2_cl_markers_res.0.1_no_MT_RP <- meso_top_cl_markers_res.0.1_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

meso_levels <- c(4,1,0,3,2)

Idents(meso) <- "RNA_css_snn_res.0.1"
levels(meso) <- meso_levels
meso$RNA_css_snn_res.0.1 <- factor(meso$RNA_css_snn_res.0.1, levels=meso_levels)

meso_top2_cl_markers_res.0.1_no_MT_RP <- meso_top2_cl_markers_res.0.1_no_MT_RP %>% group_by(cluster)  %>% arrange(factor(cluster, levels=meso_levels))

avg_cond_cl <- sapply(levels(meso$RNA_css_snn_res.0.1), function(x) rowMeans(meso@assays$RNA@data[,which(meso$RNA_css_snn_res.0.1==x)]))
# scale the data as 0 to 1 (instead)
# for top 2 markers
mat_norm <- apply(avg_cond_cl[meso_top2_cl_markers_res.0.1_no_MT_RP$gene,], 1, function(x) (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)))
col.order <- meso_top2_cl_markers_res.0.1_no_MT_RP$gene
mat_norm <- mat_norm[,col.order]

pdf("Plots/Meso_heatmap_scaled_top2_markers_cl_res.0.1.pdf", width = 5, height = 9)
heatmap.2(t(mat_norm), Colv=NA, Rowv = NA, trace="none", scale="none", margins=c(7,9), cexRow=1, cexCol=1, col =  colorRampPalette(c("white", "black"))(91), key=T, keysize = 2, density.info = "none") 
dev.off()

```

```{r plot mesoderm cells}

avg_av_cond_cl_meso <- sapply(levels(meso$RNA_css_snn_res.0.1), function(x) rowMeans(meso@assays$RNA@data[,which(meso$RNA_css_snn_res.0.1==x)]))
hcl_cl_meso <- hclust(as.dist(1-qlcMatrix::corSparse(avg_av_cond_cl_meso[VariableFeatures(meso),])), method="ward.D2")
# cols_cl_meso <- setNames(colorRampPalette(c("#0A4A3D", "#0E6B59", "#138D75", "#43A995", "#73C6B6"))(length(levels(meso$RNA_css_snn_res.0.1))), levels(meso$RNA_css_snn_res.0.1)[hcl_cl_meso$order])

```


```{r cell proportions within mesoderm clusters}

pdf("Plots/Meso_cell_prop_RNA_css_snn_res.0.1_orig.ident.pdf", width = 8, height = 7)
df <- as.data.frame(table(meso$RNA_css_snn_res.0.1, meso$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.1), y=Freq, x=orig.ident)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mesoderm cells") +
    scale_fill_manual(values= c("#0A4A3D", "#0E6B59", "#138D75", "#43A995", "#73C6B6")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Meso_cell_prop_RNA_css_snn_res.0.1_sample_timepoint.pdf", width = 8, height = 7)
df <- as.data.frame(table(meso$RNA_css_snn_res.0.1, meso$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.1), y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mesoderm cells") +
    scale_fill_manual(values= c("#0A4A3D", "#0E6B59", "#138D75", "#43A995", "#73C6B6")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Meso_cell_prop_orig.ident_RNA_css_snn_res.0.1.pdf", width = 8, height = 7)
df <- as.data.frame(table(meso$RNA_css_snn_res.0.1, meso$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.1"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=orig.ident, y=Freq, x=RNA_css_snn_res.0.1)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mesoderm cells") +
    scale_fill_manual(values= col_sample) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Meso_cell_prop_sample_timepoint_cell_phase.pdf", width = 6, height = 7)
df <- as.data.frame(table(meso$Phase, meso$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "Phase"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=Phase, y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample per timepoint") + 
    ylab("Proportion of mesoderm cells") +
    scale_fill_manual(values= c("#bdbdbd", "#737373",  "#252525")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()

```
#### Perform wicoxon test on all mesoderm clusters RNA_css_snn_res.0.1
```{r wicoxon test all mesoderm clusters separately RNA_css_snn_res.0.1}

Idents(meso) <- "RNA_css_snn_res.0.1"

# Split meso object into a list based on the cluster in RNA_css_snn_res.0.1
meso.clusters <- SplitObject(meso, split.by = "RNA_css_snn_res.0.1")

# Generate a list of data frames with DEGs
meso_de_tab <- lapply(meso.clusters, function(cl){
  de_tab <- wilcoxauc(cl, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
  blacklist <- c(grep("^MT-", de_tab$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
  diff_genes <- setdiff(de_tab$feature,blacklist)
  de_tab <- de_tab %>% filter(feature %in% diff_genes)
})


# Add number of DEGs 
meso_no.up <- c()
meso_no.down <- c()
for (i in meso_de_tab) {
  meso_no.up <- rbind(meso_no.up, sum(i$outcome == "Up-regulated"))
  meso_no.down <- rbind(meso_no.down, sum(i$outcome ==  "Down-regulated"))
}

# Createa a table with no. DEGs
meso_no.DEGs <- matrix(c("0","1","4","2","3"), ncol=1, byrow=TRUE)
meso_no.DEGs <- as.data.frame(meso_no.DEGs)
colnames(meso_no.DEGs) <- c("cluster")

meso_no.DEGs$no.up <- meso_no.up
meso_no.DEGs$no.down <- meso_no.down
meso_no.DEGs$no.total <- meso_no.down + meso_no.up
meso$RNA_css_snn_res.0.1 <- factor(meso$RNA_css_snn_res.0.1, levels=c(0,1,4,2,3)) # Change the order just for ease addition to the table
meso_no.DEGs$no.cells <- as.integer(table(meso$RNA_css_snn_res.0.1))
meso$RNA_css_snn_res.0.1 <- factor(meso$RNA_css_snn_res.0.1, levels=c(1,0,2,3,4))  # Return to the original order for future use
meso_no.DEGs <- meso_no.DEGs[order(-meso_no.DEGs$no.total),]
meso_no.DEGs$cluster <- factor(meso_no.DEGs$cluster,levels =meso_no.DEGs$cluster)
#3,1,0,2,4

# Define colors
col_order_degs_meso <- c("#0E6B59","#73C6B6","#43A995","#138D75","#0A4A3D") #3,1,0,2,4

# Plot 
pdf("Plots/Meso_lollipop_DEGs_up.pdf", height = 7, width = 5)
ggplot(meso_no.DEGs, aes(x = cluster, y = no.up)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.up), col = col_order_degs_meso) +
  geom_point(aes(x=cluster, y=no.up), size = 8, col = col_order_degs_meso) +
  geom_text(aes(x=cluster,  y=no.up, label = no.up), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of up-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/Meso_lollipop_DEGs_down.pdf", height = 7, width = 5)
ggplot(meso_no.DEGs, aes(x = cluster, y = no.down)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.down), col = col_order_degs_meso) +
  geom_point(aes(x=cluster, y=no.down), size = 8, col = col_order_degs_meso) +
  geom_text(aes(x=cluster,  y=no.down, label = no.down), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of down-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/Meso_lollipop_DEGs_total.pdf", height = 3, width = 4)
ggplot(meso_no.DEGs, aes(x = cluster), y = no.total) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.total), col = "gray50") +
  geom_point(aes(x=cluster, y=no.total, size = no.cells), col = col_order_degs_meso) +
  #geom_text(aes(x=cluster,  y=no.total, label = no.total), color = "black", size = 3) +
  theme_classic() +
  ylim(0,800) +
  xlab("Cluster") + 
  ylab("Number of DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()


# Volcano plots
pdf("Plots/Meso_volcano_DE_KO_vs_WT_all.pdf", height = 5, width = 8) # Cluster order is 0,1,4,2,3
lapply(meso_de_tab, function(cl) {ggplot(cl, aes(x = logFC, y = -log10(padj),label=DEG)) +
  # geom_point(shape=21, size=2, aes(fill=factor(outcome)), position = "jitter") +
  geom_point(colour = "grey", alpha = 0.5) +
  geom_point(data =  filter(cl, outcome == "Up-regulated" & logFC > log(1.1)), size = 1,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(cl, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
})
dev.off()


write.csv(meso_de_tab, "Plots/Meso_DE_wilcoxauc_all.csv") # Cluster order is 0,1,4,2,3
# sapply(names(de_tab), 
#  function (x) write.csv(de_tab[[x]], file=paste("Plots/Meso_DE_wilcoxauc_",x, ".csv")))


## Find total DEGs between mesoderm cells
meso_de_all <- wilcoxauc(meso, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
blacklist <- c(grep("^MT-", meso_de_all$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(meso_de_all$feature,blacklist)
meso_de_all <- meso_de_all %>% filter(feature %in% diff_genes)
meso_de_all$padj[meso_de_all$padj == 0.000000e+00] <- min(meso_de_all$padj[which(meso_de_all$padj != 0)])

pdf("Plots/Meso_volcano_DE_KO_vs_WT.pdf", height = 5, width = 8)
ggplot(meso_de_all, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(meso_de_all, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(meso_de_all, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

```
```{r wicoxon test on mesoderm clusters 3,1,0,2 RNA_css_snn_res.0.1}

Idents(meso) <- "RNA_css_snn_res.0.1"
meso_3_1_0_2 <- subset(meso, idents = c(3,1,0,2))
Idents(meso_3_1_0_2) <- "condition"

# Run wilcoxon test
DE_wilcoxauc_meso_3_1_0_2 <- wilcoxauc(meso_3_1_0_2, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_meso_3_1_0_2 <- DE_wilcoxauc_meso_3_1_0_2 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_meso_3_1_0_2) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_meso_3_1_0_2$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(DE_wilcoxauc_meso_3_1_0_2$feature,blacklist)
DE_wilcoxauc_meso_3_1_0_2 <- DE_wilcoxauc_meso_3_1_0_2 %>% filter(feature %in% diff_genes)
DE_wilcoxauc_meso_3_1_0_2$padj[DE_wilcoxauc_meso_3_1_0_2$padj == 0.000000e+00] <- min(DE_wilcoxauc_meso_3_1_0_2$padj[which(DE_wilcoxauc_meso_3_1_0_2$padj != 0)])

pdf("Plots/Meso_volcano_DE_KO_vs_WT_cl3_1_0_2.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_meso_3_1_0_2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_meso_3_1_0_2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_meso_3_1_0_2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/Meso_volcano_DE_KO_vs_WT_cl3_1_0_2_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_meso_3_1_0_2, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_meso_3_1_0_2, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_meso_3_1_0_2, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  # geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_meso_3_1_0_2, "Plots/Meso_volcano_DE_KO_vs_WT_cl3_1_0_2.csv") 

```
```{r feature plots mesoderm}

# Markers based on Loh et al (2016) Cell
meso_markers <- c("HAND1","FOXF1","ISL1","NKX2-5", "MEF2C", "GATA4","GATA6", # lateral/cardiac mesoderm 
                  "TBX6","DLL3", "EVX1", "CDX2", # paraxial mesoderm (and anterior primitive streak)
                  "MESP1","MESP2",  # mid/posterior primitive streak
                  "GSC","EOMES", # mid/posterior primitive streak
                  "BST2", "SST", "IGFBP3", # genes chosen from DE analysis - DOWN in KO
                  "FN1", "SERPINE2", "NBDY","IGFBP7","TGFBI", "SOX4" # genes chosen from DE analysis - UP in KO
                  )

png("Plots/Meso_featureplots_markers.png", width = 7000, height = 4000, res = 500)
plotMultiFeatures(Embeddings(meso, "umap_css"),
                  meso@assays$RNA@data[meso_markers,],
                  ncol=6, colorPal=blue_colscheme, cex=0.5, par_cex = 0.8, mar = c(1,1,2,2), cex.main=1, sort_by_value=T)
dev.off()

```

# Mural cells only

#### Find mural cluster markers
```{r mural cluster markers res.0.2}

Idents(mural) <- "RNA_css_snn_res.0.2"

table(mural$orig.ident, mural$RNA_css_snn_res.0.2)

cl_markers_res.0.2 <- FindAllMarkers(mural, min.pct = 0.25, logfc.threshold = log(1.2))
   
top2_cl_markers_res.0.2 <- cl_markers_res.0.2 %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
top3_cl_markers_res.0.2 <- cl_markers_res.0.2 %>% group_by(cluster) %>%  slice_max(n = 3, order_by = avg_log2FC)
top10_cl_markers_res.0.2 <- cl_markers_res.0.2 %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.2 <- cl_markers_res.0.2 %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.2 <- cl_markers_res.0.2 %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.2, "Plots/Mural_top_cl_markers_res.0.2.csv")
write.csv(top50_cl_markers_res.0.2, "Plots/Mural_top50_cl_markers_res.0.2.csv")

# top_cl_markers_unique <- top3_cl_markers_res.0.2[!duplicated(top3_cl_markers_res.0.2[ , "gene"]), ]  # Delete rows
# top2_cl_markers_unique <- top_cl_markers_unique %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

# Remove mitochondria and ribosome genes from the list
blacklist <- c(grep("^MT-", cl_markers_res.0.2$gene, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(cl_markers_res.0.2$gene, blacklist)
cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2 %>% filter(gene %in% diff_genes)
nrow(cl_markers_res.0.2_no_MT_RP)

# Redo top marker selection
top2_cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC) 
top5_cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 5, order_by = avg_log2FC) 
top10_cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 10, order_by = avg_log2FC)
top50_cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>%  slice_max(n = 50, order_by = avg_log2FC)

top_cl_markers_res.0.2_no_MT_RP <- cl_markers_res.0.2_no_MT_RP %>%
    group_by(cluster) %>%
     arrange(desc(avg_log2FC), .by_group = TRUE)

write.csv(top_cl_markers_res.0.2_no_MT_RP, "Plots/Mural_top_cl_markers_res.0.2_no_MT_RP.csv")
write.csv(top50_cl_markers_res.0.2_no_MT_RP, "Plots/Mural_top50_cl_markers_res.0.2_no_MT_RP.csv")

```

#### Plot mural cluster markers

```{r heatmap mural top markers by cluster res.0.2}

mural_top_cl_markers_res.0.2_no_MT_RP <- read.csv("Plots/Mural_cells/Mural_top50_cl_markers_res.0.2_no_MT_RP.csv")
mural_top2_cl_markers_res.0.2_no_MT_RP <- mural_top_cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)

mural_levels <- c(2,1,3,4,0)

Idents(mural) <- "RNA_css_snn_res.0.1"
levels(mural) <- mural_levels
mural$RNA_css_snn_res.0.2 <- factor(mural$RNA_css_snn_res.0.2, levels=mural_levels)

mural_top2_cl_markers_res.0.2_no_MT_RP <- mural_top2_cl_markers_res.0.2_no_MT_RP %>% group_by(cluster) %>% arrange(factor(cluster, levels=mural_levels))

avg_cond_cl <- sapply(levels(mural$RNA_css_snn_res.0.2), function(x) rowMeans(mural@assays$RNA@data[,which(mural$RNA_css_snn_res.0.2==x)]))
# scale the data as 0 to 1 (instead)
# for top 2 markers
mat_norm <- apply(avg_cond_cl[mural_top2_cl_markers_res.0.2_no_MT_RP$gene,], 1, function(x) (x-min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)))
col.order <- mural_top2_cl_markers_res.0.2_no_MT_RP$gene
mat_norm <- mat_norm[,col.order]

pdf("Plots/Mural_heatmap_scaled_top2_markers_cl_res.0.2.pdf", width = 5, height = 9)
heatmap.2(t(mat_norm), Colv=NA, Rowv = NA, trace="none", scale="none", margins=c(7,9), cexRow=1, cexCol=1, col =  colorRampPalette(c("white", "black"))(91), key=T, keysize = 2, density.info = "none") 
dev.off()

```

```{r plot mural}

avg_av_cond_cl_mural <- sapply(levels(mural$RNA_css_snn_res.0.2), function(x) rowMeans(mural@assays$RNA@data[,which(mural$RNA_css_snn_res.0.2==x)]))
hcl_cl_mural <- hclust(as.dist(1-qlcMatrix::corSparse(avg_av_cond_cl_mural[VariableFeatures(mural),])), method="ward.D2")
# cols_cl_mural <- setNames(colorRampPalette(c("#F7DC6F", "#F2CA27", "#D3AC0D", "#AB8B0A", "#876E08"))(length(levels(mural$RNA_css_snn_res.0.2))), levels(mural$RNA_css_snn_res.0.2)[hcl_cl_mural$order])

```


```{r cell proportions within mural clusters}

pdf("Plots/Mural_cell_prop_RNA_css_snn_res.0.2_orig.ident.pdf", width = 8, height = 7)
df <- as.data.frame(table(mural$RNA_css_snn_res.0.2, mural$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.2"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.2), y=Freq, x=orig.ident)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mural cells") +
    scale_fill_manual(values= c("#F7DC6F", "#F2CA27", "#D3AC0D", "#AB8B0A", "#876E08")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Mural_cell_prop_RNA_css_snn_res.0.2_sample_timepoint.pdf", width = 8, height = 7)
df <- as.data.frame(table(mural$RNA_css_snn_res.0.2, mural$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.2"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=factor(RNA_css_snn_res.0.2), y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mural cells") +
    scale_fill_manual(values= c("#F7DC6F", "#F2CA27", "#D3AC0D", "#AB8B0A", "#876E08")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Mural_cell_prop_orig.ident_RNA_css_snn_res.0.2.pdf", width = 8, height = 7)
df <- as.data.frame(table(mural$RNA_css_snn_res.0.2, mural$orig.ident))
colnames(df)[which(names(df) == "Var1")] <- "RNA_css_snn_res.0.2"
colnames(df)[which(names(df) == "Var2")] <- "orig.ident"

ggplot(df, aes(fill=orig.ident, y=Freq, x=RNA_css_snn_res.0.2)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample") + 
    ylab("Proportion of mural cells") +
    scale_fill_manual(values= col_sample) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()


pdf("Plots/Mural_cell_prop_sample_timepoint_cell_phase.pdf", width = 6, height = 7)
df <- as.data.frame(table(mural$Phase, mural$sample_timepoint))
colnames(df)[which(names(df) == "Var1")] <- "Phase"
colnames(df)[which(names(df) == "Var2")] <- "sample_timepoint"

ggplot(df, aes(fill=Phase, y=Freq, x=sample_timepoint)) + 
    geom_bar(position="fill", stat="identity") + 
    ggplot2::theme_classic() + 
    xlab("Sample per timepoint") + 
    ylab("Proportion of mural cells") +
    scale_fill_manual(values= c("#bdbdbd", "#737373",  "#252525")) +
    theme(axis.title.x = element_text(size = 16), axis.text.x = element_text(size = 13, angle = 45, hjust = 1), axis.title.y = element_text(size = 16), axis.text.y = element_text(size = 16))  + 
    labs(fill='Cluster') 
dev.off()

```

#### Perform wilcoxon test on all mural clusters RNA_css_snn_res.0.2
```{r wilcoxon test all mesoderm clusters separately RNA_css_snn_res.0.2}

Idents(mural) <- "RNA_css_snn_res.0.2"
# Split mural object into a list based on the cluster in RNA_css_snn_res.0.2
mural.clusters <- SplitObject(mural, split.by = "RNA_css_snn_res.0.2")

# Generate a list of data frames with DEGs
mural_de_tab <- lapply(mural.clusters, function(cl){
  de_tab <- wilcoxauc(cl, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
  blacklist <- c(grep("^MT-", de_tab$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
  diff_genes <- setdiff(de_tab$feature,blacklist)
  de_tab <- de_tab %>% filter(feature %in% diff_genes)
  })

mural_de_tab <- lapply(mural_de_tab, function(df){
  df$padj[df$padj == 0.000000e+00] <- min(df$padj[which(df$padj != 0)])
  return(df)
})


# Add number of DEGs 
mural_no.up <- c()
mural_no.down <- c()
for (i in de_tab) {
  mural_no.up <- rbind(mural_no.up, sum(i$outcome == "Up-regulated"))
  mural_no.down <- rbind(mural_no.down, sum(i$outcome ==  "Down-regulated"))
}

# Createa a table with no. DEGs
mural_no.DEGs <- matrix(c("3","1","0","4","2"), ncol=1, byrow=TRUE)
mural_no.DEGs <- as.data.frame(mural_no.DEGs)
colnames(mural_no.DEGs) <- c("cluster")

mural_no.DEGs$no.up <- mural_no.up
mural_no.DEGs$no.down <- mural_no.down
mural_no.DEGs$no.total <- mural_no.down + mural_no.up
mural$RNA_css_snn_res.0.2 <- factor(mural$RNA_css_snn_res.0.2, levels=c(3,1,0,4,2)) # Change the order just for ease addition to the table
mural_no.DEGs$no.cells <- as.integer(table(mural$RNA_css_snn_res.0.2))
mural$RNA_css_snn_res.0.2 <- factor(mural$RNA_css_snn_res.0.2, levels=c(4,1,0,2,3))  # Return to the original order for future use
mural_no.DEGs <- mural_no.DEGs[order(-mural_no.DEGs$no.total),]
mural_no.DEGs$cluster <- factor(mural_no.DEGs$cluster,levels = mural_no.DEGs$cluster)
#2,1,3,4,0

# Define colors
col_order_degs_mural <- c( "#AB8B0A", "#F2CA27", "#876E08", "#F7DC6F", "#D3AC0D") #2,1,3,4,0

# Plot 
pdf("Plots/Mural_lollipop_DEGs_up.pdf", height = 7, width = 5)
ggplot(mural_no.DEGs, aes(x = cluster, y = no.up)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.up), col = col_order_degs_mural) +
  geom_point(aes(x=cluster, y=no.up), size = 8, col = col_order_degs_mural) +
  geom_text(aes(x=cluster,  y=no.up, label = no.up), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of up-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/Mural_lollipop_DEGs_down.pdf", height = 7, width = 5)
ggplot(mural_no.DEGs, aes(x = cluster, y = no.down)) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.down), col = col_order_degs_mural) +
  geom_point(aes(x=cluster, y=no.down), size = 8, col = col_order_degs_mural) +
  geom_text(aes(x=cluster,  y=no.down, label = no.down), color = "black", size = 3) +
  theme_classic() +
  xlab("Cluster") + 
  ylab("Number of down-regulated DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()

pdf("Plots/Mural_lollipop_DEGs_total.pdf", height = 3, width = 4)
ggplot(mural_no.DEGs, aes(x = cluster), y = no.total) +
  geom_segment(aes(x = cluster, xend = cluster, y = 0, yend = no.total), col = "gray50") +
  geom_point(aes(x=cluster, y=no.total, size = no.cells), col = col_order_degs_mural) +
  #geom_text(aes(x=cluster,  y=no.total, label = no.total), color = "black", size = 3) +
  theme_classic() +
  ylim(0,750) +
  xlab("Cluster") + 
  ylab("Number of DEGs") +
  theme(axis.text.x = element_text( angle = 45, hjust = 1)) 
dev.off()


# Volcano plots
pdf("Plots/Mural_volcano_DE_KO_vs_WT_all.pdf", height = 5, width = 8) # Cluster order is 3,1,0,4,2
lapply(mural_de_tab, function(cl) {ggplot(cl, aes(x = logFC, y = -log10(padj),label=DEG)) +
  # geom_point(shape=21, size=2, aes(fill=factor(outcome)), position = "jitter") +
  geom_point(colour = "grey", alpha = 0.5) +
  geom_point(data =  filter(cl, outcome == "Up-regulated" & logFC > log(1.1)), size = 1,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(cl, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj")) 
})
dev.off()


write.csv(mural_de_tab, "Plots/Mural_DE_wilcoxauc_all.csv") # Cluster order is 3,1,0,4,2
# sapply(names(de_tab), 
#  function (x) write.csv(de_tab[[x]], file=paste("Plots/Mural_DE_wilcoxauc_",x, ".csv")))



## Find total DEGs between mural cells
mural_de_all <- wilcoxauc(mural, group_by = "condition") %>%
    filter(group == "KO") %>%
    mutate(DE = abs(logFC)>log(1.1) & padj < 0.01) %>%
    mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))%>% 
    mutate(outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated", logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated", TRUE ~ "Unchanged"))
blacklist <- c(grep("^MT-", mural_de_all$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(mural_de_all$feature,blacklist)
mural_de_all <- mural_de_all %>% filter(feature %in% diff_genes)
mural_de_all$mural_de_all[mural_de_all$padj == 0.000000e+00] <- min(mural_de_all$padj[which(mural_de_all$padj != 0)])

pdf("Plots/Mural_volcano_DE_KO_vs_WT.pdf", height = 5, width = 8)
ggplot(mural_de_all, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray",alpha = 0.5, size = 1) +
  geom_point(data =  filter(mural_de_all, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(mural_de_all, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

```

```{r wicoxon test on mural clusters 2,1,3,4 RNA_css_snn_res.0.2}

Idents(mural) <- "RNA_css_snn_res.0.2"
mural_2_1_3_4 <- subset(mural, idents = c(2,1,3,4))
Idents(mural_2_1_3_4) <- "condition"

# Run wilcoxon test
DE_wilcoxauc_mural_2_1_3_4 <- wilcoxauc(mural_2_1_3_4, group_by = "condition") %>%
  filter(group == "KO") %>%
  mutate(DE = abs(logFC)>log(1.1) & padj < 0.1) %>%
  mutate(DEG = ifelse(abs(logFC)>log(1.1) & padj < 0.01, feature, NA))

# Add to the table another column wtih outcome
DE_wilcoxauc_mural_2_1_3_4 <- DE_wilcoxauc_mural_2_1_3_4 %>% 
  mutate(
    outcome = case_when(logFC > log(1.1) & padj < 0.01 ~ "Up-regulated",
                           logFC < -log(1.1) & padj < 0.01 ~ "Down-regulated",
                           TRUE ~ "Unchanged")
    )
head(DE_wilcoxauc_mural_2_1_3_4) 

# Remove mitochondria and ribosomal protein coding genes from the DE genes
blacklist <- c(grep("^MT-", DE_wilcoxauc_mural_2_1_3_4$feature, value=T), read.table("/home/marinani/Scripts/Databases/Gene_lists/RPgenes_bak.txt")[,1])
diff_genes <- setdiff(DE_wilcoxauc_mural_2_1_3_4$feature,blacklist)
DE_wilcoxauc_mural_2_1_3_4 <- DE_wilcoxauc_mural_2_1_3_4 %>% filter(feature %in% diff_genes)
DE_wilcoxauc_mural_2_1_3_4$padj[DE_wilcoxauc_mural_2_1_3_4$padj == 0.000000e+00] <- min(DE_wilcoxauc_mural_2_1_3_4$padj[which(DE_wilcoxauc_mural_2_1_3_4$padj != 0)])

pdf("Plots/Mural_volcano_DE_KO_vs_WT_cl2_1_3_4.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_mural_2_1_3_4, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_mural_2_1_3_4, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_mural_2_1_3_4, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

pdf("Plots/Mural_volcano_DE_KO_vs_WT_cl2_1_3_4_no_label.pdf", height = 5, width = 8)
ggplot(DE_wilcoxauc_mural_2_1_3_4, aes(x = logFC, y = -log10(padj),label=DEG)) +
  geom_point(colour = "grey", fill= "gray", alpha = 0.5, size = 1) +
  geom_point(data =  filter(DE_wilcoxauc_mural_2_1_3_4, outcome == "Up-regulated" & logFC > log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black") +
  geom_point(data =  filter(DE_wilcoxauc_mural_2_1_3_4, outcome == "Down-regulated" & logFC < -log(1.1)), size = 1.5,shape = 21,fill = "black",colour = "black")+
  # geom_text_repel(max.overlaps = 50) +
  theme_classic() + 
  geom_vline(xintercept=c(0),  linetype="dotted") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"p_adj"))
dev.off()

write.csv(DE_wilcoxauc_mural_2_1_3_4, "Plots/Mural_volcano_DE_KO_vs_WT_cl2_1_3_4.csv") 

```

```{r feature plots mural}

mural_markers <- c("KCNJ8","DCN","RGS5","CSPG4","ACTA2",  "LUM", # general markers
                  "TAC1", "COL18A1", "VIM", "TPM1", "IGFBP7", "DLK1","PDGFRB","TPM2","MYL6","MEF2C", # genes chosen from DE analysis - DOWN in KO
                  "IGFBP5", "PDGFRA", "SELENOP","NR2F1","MEIS2","SOX4","CXCL14","MEIS1" # genes chosen from DE analysis - UP in KO
                  )

png("Plots/Mural_featureplots_markers.png", width = 7000, height = 4000, res = 500)
plotMultiFeatures(Embeddings(mural, "umap_css"),
                  mural@assays$RNA@data[mural_markers,],
                  ncol=6, colorPal=blue_colscheme, cex=0.5, par_cex = 0.8, mar = c(1,1,2,2), cex.main=1, sort_by_value=T)
dev.off()

```

# Differential abundace with Milo

## MiloR on all cells

Milo is a tool for analysis of complex single cell datasets generated from replicated multi-condition experiments, which detects changes in composition between conditions. While differential abundance (DA) is commonly quantified in discrete cell clusters, Milo uses partally overlapping neighbourhoods of cells on a KNN graph. Starting from a graph that faithfully recapitulates the biology of the cell population, Milo analysis consists of 3 steps:

Sampling of representative neighbourhoods
Testing for differential abundance of conditions in all neighbourhoods
Accounting for multiple hypothesis testing using a weighted FDR procedure that accounts for the overlap of neighbourhoods

#### Prepare environment and objects
```{r prepare the environment}

# BiocManager::install("miloR")
# citation("miloR")

library(miloR)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(patchwork)

```

Convert the seurat object into a single cell experiemtn object. Then convert the SCE further to an extended SCE object class that can store the information about neighborhoods on the kNN graph, a Milo object.
```{r convert Seurat object into SCE and into Milo objects}

Idents(seurat) <- "condition"
seurat_sce <- as.SingleCellExperiment(seurat)
seurat_milo <- Milo(seurat_sce)

```

Generate a kNN graph.
This is stored in the graph slot, in igraph format. The miloR package includes functionality to build and store the graph from the PCA dimensions stored in the reducedDim slot.
```{r kNN graph construction}

seurat_milo <- buildGraph(seurat_milo, k = 50, d = 30)

```

#### Define representative neighborhoods sl

The neighborhood of a cells i defined as the group of cells connected by an edge in the kNN graph ot the index cell. 
For efficiency, Milo doesn't test for DA in the neighborhood of every cell, but samples as indices a subset of representative cells, using a kNN sampling algorithm used by Gut et al (2015).
```{r sampling}

# For sampling you need to define a few parameters:
# - prop: the proportion of cells to randomly sample to start with (usually 0.1 - 0.2 is sufficient)
# - k: the k to use for KNN refinement (we recommend using the same k used for KNN graph building)
# - d: the number of reduced dimensions to use for KNN refinement (we recommend using the same d used for KNN graph building)
# - refined indicated whether you want to use the sampling refinement algorith, or just pick cells at random. The default and recommended way to go is to use refinement. The only situation in which you might consider using random instead, is if you have batch corrected your data with a graph based correction algorithm, such as BBKNN, but the results of DA testing will be suboptimal.

seurat_milo <- makeNhoods(seurat_milo, prop = 0.1, k = 50, d=30, refined = TRUE)

# Check the size of neighborhoods as this affects the power of DA testing
plotNhoodSizeHist(seurat_milo)
# Empirically, the authors found its best to have a distribution peaking between 50 and 100, or at least to have a distribution peaking above 20. 
# Otherwise you might consider rerunning makeNhoods increasing k and/or prop.

```


Count how many cells from each sample are in each neighbourhood.
Milo leverages the variation in cell numbers between replicates for the same experimental condition to test for differential abundance. Therefore we have to count how many cells from each sample are in each neighbourhood. We need to use the cell metadata and specify which column contains the sample information.

```{r count cells in neighborhoods}

seurat_milo <- countCells(seurat_milo, meta.data = data.frame(colData(seurat_milo)), sample="orig.ident")

# This adds to the Milo object a  matrix, where  is the number of neighbourhoods and  is the number of experimental samples. Values indicate the number of cells from each sample counted in a neighbourhood. This count matrix will be used for DA testing.
head(nhoodCounts(seurat_milo))

```


#### Differential abundance testing

Now we are all set to test for differential abundance in neighbourhoods. We implement this hypothesis testing in a generalized linear model (GLM) framework, specifically using the Negative Binomial GLM implementation in edgeR.

We first need to think about our experimental design. The design matrix should match each sample to the experimental condition of interest for DA testing.

```{r define experimental design}

seurat_design <- data.frame(colData(seurat_milo))[,c("orig.ident", "condition", "sample","timepoint", "replicate","KO_replicate")]
## Convert batch info from integer to factor
seurat_design$replicate <- as.factor(seurat_design$replicate)
seurat_design <- distinct(seurat_design)
rownames(seurat_design) <- seurat_design$orig.ident

seurat_design

```


```{r DA testing}

seurat_milo <- calcNhoodDistance(seurat_milo, d=30)

seurat_design$condition <- factor(seurat_design$condition, levels = c("WT","KO"))
da_results <- testNhoods(seurat_milo, design = ~ timepoint + condition, design.df = seurat_design)

da_results %>%
  arrange(- SpatialFDR) %>%
  head()

# Inspect the distribution of uncorrected P values, to verify that the test was balanced
pdf("Plots/histo_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 10, height = 7)
ggplot(da_results, aes(PValue)) + geom_histogram(bins=50) + theme_classic()
dev.off()

# Visualize the test results with a volcano plot (each point here represents a neighbourhood, not a cell)
pdf("Plots/volcano_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 5, height = 3.5)
ggplot(da_results, aes(logFC, -log10(SpatialFDR))) +
  geom_point(colour = "grey", alpha = 0.5, size = 0.8) +
  geom_point(data = filter(da_results, -log10(SpatialFDR) > 1), size = 1.2,shape = 21,fill = "black",colour = "black") + 
  theme_classic()
  # geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR)
dev.off()
  
```


To visualize DA results relating them to the embedding of single cells, we can build an abstracted graph of neighbourhoods that we can superimpose on the single-cell embedding. Here each node represents a neighbourhood, while edges indicate how many cells two neighbourhoods have in common. Here the layout of nodes is determined by the position of the index cell in the UMAP embedding of all single-cells. The neighbourhoods displaying singificant DA are colored by their log-Fold Change.

```{r visualize}

seurat_milo <- buildNhoodGraph(seurat_milo, overlap = 80)

## Plot neighbourhood graph
nh_graph_pl <- plotNhoodGraphDA(seurat_milo, da_results, layout="UMAP_CSS",alpha=0.1) 

da_results <- annotateNhoods(seurat_milo, da_results, coldata_col = "RNA_css_snn_res.0.4")
head(da_results)

nh_graph_pl$data <- nh_graph_pl$data %>% arrange(abs(colour_by))

# While neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.
ggplot(da_results, aes(RNA_css_snn_res.0.4_fraction)) + geom_histogram(bins=50)
da_results$RNA_css_snn_res.0.4 <- ifelse(da_results$RNA_css_snn_res.0.4_fraction < 0.5, "Overlapping", da_results$RNA_css_snn_res.0.4)


pdf("Plots/plotDAbeeswarm_miloR_timepoint_condition_k50_prop0.1.pdf", width = 4, height = 8)
plotDAbeeswarm(da_results, group.by = "RNA_css_snn_res.0.4") + theme_classic() +scale_color_gradient2(high = c("#F9B99D","#B5202D"), low = c("#2468AD","#9EC6DF"))
dev.off()

pdf("Plots/umap_miloR_timepoint_condition_k50_prop0.1.pdf", width = 8, height = 4.5)
nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

pdf("Plots/umap_miloR_by_neighborhood_k50_prop0.1.pdf", width = 6, height = 4.5)
plotNhoodGraph(seurat_milo, colour_by =  "RNA_css_snn_res.0.4", layout="UMAP_CSS") 
dev.off()

png("Plots/umap_miloR_timepoint_condition_k50_prop0.1.png", width = 2700, height = 1750, res = 500)
nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

png("Plots/umap_miloR_timepoint_condition_org_colscheme_k50_prop0.1.png", width = 2700, height = 1750, res = 500)
nh_graph_pl + plot_layout(guides="collect") 
dev.off()

```

```{r save miloR on all cells}

saveRDS(seurat_milo, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/seurat_milo_k50_prop0.1.rds")
saveRDS(da_results, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/DA_results_milo_k50_prop0.1.rds")
write.csv(da_results,"/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Plots/DA_results_milo_k50_prop0.1.csv")

```



## MiloR on endothelial cells

#### Prepare environment and objects
```{r convert Seurat EC object into SCE and into Milo objects}

ec_sce <- as.SingleCellExperiment(ec)
ec_milo <- Milo(ec_sce)

```

```{r kNN graph construction on ECs}

ec_milo <- buildGraph(ec_milo, k = 50, d = 30)

```

#### Define representative neighborhoods sl

```{r sampling ECs}

ec_milo <- makeNhoods(ec_milo, prop = 0.1, k = 50, d=30, refined = TRUE)

# Check the size of neighborhoods as this affects the power of DA testing
plotNhoodSizeHist(ec_milo)
# Empirically, the authors found its best to have a distribution peaking between 50 and 100, or at least to have a distribution peaking above 20. 
# Otherwise you might consider rerunning makeNhoods increasing k and/or prop.

```

```{r count cells in EC neighborhoods}

ec_milo <- countCells(ec_milo, meta.data = data.frame(colData(ec_milo)), sample="orig.ident")

# This adds to the Milo object a  matrix, where  is the number of neighbourhoods and  is the number of experimental samples. Values indicate the number of cells from each sample counted in a neighbourhood. This count matrix will be used for DA testing.
head(nhoodCounts(ec_milo))

```


#### Differential abundance testing

```{r define experimental design for EC milo}

ec_design <- data.frame(colData(ec_milo))[,c("orig.ident", "condition", "sample","timepoint")]
## Convert batch info from integer to factor
ec_design <- distinct(ec_design)
rownames(ec_design) <- ec_design$orig.ident

ec_design

```

`
```{r EC DA testing}

ec_milo <- calcNhoodDistance(ec_milo, d=30)
ec_design$condition <- factor(ec_design$condition, levels = c("WT","KO"))
ec_da_results <- testNhoods(ec_milo, design = ~ timepoint + condition, design.df = ec_design)

ec_da_results %>%
  arrange(- SpatialFDR) %>%
  head()

# Inspect the distribution of uncorrected P values, to verify that the test was balanced
pdf("Plots/EC_histo_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 10, height = 7)
ggplot(ec_da_results, aes(PValue)) + geom_histogram(bins=50) + theme_classic()
dev.off()

# Visualize the test results with a volcano plot (each point here represents a neighbourhood, not a cell)
pdf("Plots/EC_volcano_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 5, height = 3.5)
ggplot(ec_da_results, aes(logFC, -log10(SpatialFDR))) +
  geom_point(colour = "grey", alpha = 0.5, size = 0.8) +
  geom_point(data = filter(ec_da_results, -log10(SpatialFDR) > 1), size = 1.2,shape = 21,fill = "black",colour = "black") + 
  theme_classic()
  # geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR)
dev.off()
  
```

```{r visualize EC milo}

ec_milo <- buildNhoodGraph(ec_milo, overlap = 80)

## Plot neighbourhood graph
ec_nh_graph_pl <- plotNhoodGraphDA(ec_milo, ec_da_results, layout="UMAP_CSS",alpha=0.1) 

ec_da_results <- annotateNhoods(ec_milo, ec_da_results, coldata_col = "RNA_css_snn_res.0.1")
head(ec_da_results)

ec_nh_graph_pl$data <- ec_nh_graph_pl$data %>% arrange(abs(colour_by))

# While neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.
ggplot(ec_da_results, aes(RNA_css_snn_res.0.1_fraction)) + geom_histogram(bins=50)
ec_da_results$RNA_css_snn_res.0.1 <- ifelse(ec_da_results$RNA_css_snn_res.0.1_fraction < 0.5, "Overlapping", ec_da_results$RNA_css_snn_res.0.1)

pdf("Plots/EC_plotDAbeeswarm_miloR_timepoint_condition_k50_prop0.1.pdf", width = 5, height = 5)
plotDAbeeswarm(ec_da_results, group.by = "RNA_css_snn_res.0.1") + theme_classic() +scale_color_gradient2(high = c("#F9B99D","#B5202D"), low = c("#2468AD","#9EC6DF"))
dev.off()

pdf("Plots/EC_umap_miloR_by_neighborhood_k50_prop0.1.pdf", width = 4, height = 2.7)
plotNhoodGraph(ec_milo, colour_by =  "RNA_css_snn_res.0.4", layout="UMAP_CSS") 
dev.off()

pdf("Plots/EC_umap_miloR_timepoint_conditionk50_prop0.1.pdf", width = 4, height = 2.7)
ec_nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

png("Plots/EC_umap_miloR_timepoint_conditionk50_prop0.1.png", width = 2300, height = 1300, res = 500)
ec_nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

```


```{r save miloR on ECs}

saveRDS(ec_milo, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/EC_seurat_milo_k50_prop0.1.rds")
saveRDS(ec_da_results, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/EC_DA_results_milo_k50_prop0.1.rds")

```

## MiloR on mural cells

#### Prepare environment and objects
```{r convert Seurat mural object into SCE and into Milo objects}

mural_sce <- as.SingleCellExperiment(mural)
mural_milo <- Milo(mural_sce)

```

```{r kNN graph construction on mural cells}

mural_milo <- buildGraph(mural_milo, k = 50, d = 30)

```

#### Define representative neighborhoods sl

```{r sampling mural cells}

mural_milo <- makeNhoods(mural_milo, prop = 0.1, k = 50, d=30, refined = TRUE)

# Check the size of neighborhoods as this affects the power of DA testing
plotNhoodSizeHist(mural_milo)
# Empirically, the authors found its best to have a distribution peaking between 50 and 100, or at least to have a distribution peaking above 20. 
# Otherwise you might consider rerunning makeNhoods increasing k and/or prop.

```

```{r count cells in mural cells neighborhoods}

mural_milo <- countCells(mural_milo, meta.data = data.frame(colData(mural_milo)), sample="orig.ident")

# This adds to the Milo object a  matrix, where  is the number of neighbourhoods and  is the number of experimental samples. Values indicate the number of cells from each sample counted in a neighbourhood. This count matrix will be used for DA testing.
head(nhoodCounts(mural_milo))

```


#### Differential abundance testing

```{r define experimental design for mural milo}

mural_design <- data.frame(colData(mural_milo))[,c("orig.ident", "condition", "sample","timepoint")]
## Convert batch info from integer to factor
mural_design <- distinct(mural_design)
rownames(mural_design) <- mural_design$orig.ident

mural_design

```

`
```{r mural DA testing}

mural_milo <- calcNhoodDistance(mural_milo, d=30)
mural_design$condition <- factor(mural_design$condition, levels = c("WT","KO"))
mural_da_results <- testNhoods(mural_milo, design = ~ timepoint + condition, design.df = mural_design)

mural_da_results %>%
  arrange(- SpatialFDR) %>%
  head()

# Inspect the distribution of uncorrected P values, to verify that the test was balanced
pdf("Plots/Mural_histo_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 10, height = 7)
ggplot(mural_da_results, aes(PValue)) + geom_histogram(bins=50) + theme_classic()
dev.off()

# Visualize the test results with a volcano plot (each point here represents a neighbourhood, not a cell)
pdf("Plots/Mural_volcano_da_results_miloR_timepoint_condition_k50_prop0.1.pdf", width = 5, height = 3.5)
ggplot(mural_da_results, aes(logFC, -log10(SpatialFDR))) +
  geom_point(colour = "grey", alpha = 0.5, size = 0.8) +
  geom_point(data = filter(mural_da_results, -log10(SpatialFDR) > 1), size = 1.2,shape = 21,fill = "black",colour = "black") + 
  theme_classic()
  # geom_hline(yintercept = 1) ## Mark significance threshold (10% FDR)
dev.off()
  
```

```{r visualize mural milo}

mural_milo <- buildNhoodGraph(mural_milo, overlap = 80)

## Plot neighbourhood graph
mural_nh_graph_pl <- plotNhoodGraphDA(mural_milo, mural_da_results, layout="UMAP_CSS",alpha=0.1) 

mural_da_results <- annotateNhoods(mural_milo, mural_da_results, coldata_col = "RNA_css_snn_res.0.2")
head(mural_da_results)

mural_nh_graph_pl$data <- mural_nh_graph_pl$data %>% arrange(abs(colour_by))

# While neighbourhoods tend to be homogeneous, we can define a threshold for celltype_fraction to exclude neighbourhoods that are a mix of cell types.
ggplot(mural_da_results, aes(RNA_css_snn_res.0.2_fraction)) + geom_histogram(bins=50)
mural_da_results$RNA_css_snn_res.0.2 <- ifelse(mural_da_results$RNA_css_snn_res.0.2_fraction < 0.5, "Overlapping", mural_da_results$RNA_css_snn_res.0.2)


pdf("Plots/Mural_plotDAbeeswarm_miloR_timepoint_condition_k50_prop0.1.pdf", width = 3.5, height = 3.5)
plotDAbeeswarm(mural_da_results, group.by = "RNA_css_snn_res.0.2") + theme_classic() +scale_color_gradient2(high = c("#F9B99D","#B5202D"), low = c("#2468AD","#9EC6DF"))
dev.off()

pdf("Plots/Mural_umap_miloR_timepoint_condition_k50_prop0.1.pdf", width = 3, height = 1.7)
mural_nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

pdf("Plots/Mural_umap_miloR_by_neighborhood_k50_prop0.1.pdf", width = 3, height = 1.7)
plotNhoodGraph(mural_milo, colour_by =  "RNA_css_snn_res.0.2", layout="UMAP_CSS") 
dev.off()

png("Plots/Mural_umap_miloR_timepoint_condition_k50_prop0.1.png", width = 2300, height = 1300, res = 500)
mural_nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

png("Plots/Mural_umap_miloR_timepoint_condition_org_colscheme_k50_prop0.1.png", width = 2300, height = 1300, res = 500)
mural_nh_graph_pl + plot_layout(guides="collect") +scale_fill_distiller(type = "div",palette = 5,direction = -1)
dev.off()

```


```{r save miloR on mural cells}

saveRDS(mural_milo, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/Mural_seurat_milo_k50_prop0.1.rds")
saveRDS(mural_da_results, "/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Objects/Mural_DA_results_milo_k50_prop0.1.rds")

```

# Compare results to CROPseq MECOM result

```{r pearson correlation MECOM KO to CROPseq MECOM}

DE_ct_prob <- readRDS("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/ms/CROPseq/CROPseq_no_I_J/res.DE-anova_perturb-prop_target_vs_dummy_ct.rds")

df_DE_ct <- do.call(rbind, lapply(names(DE_ct_prob), function(ct)
  do.call(rbind, lapply(DE_ct_prob[[ct]], function(x)
    data.frame(celltype = ct, gene = rownames(x), x[,c(1:5,8)], padj = p.adjust(x$p_ANOVA, method="bonferroni"))))))
rownames(df_DE_ct) <- NULL

num_DEGs <- sapply(split(df_DE_ct,df_DE_ct$celltype), function(x) tapply(x$padj < 0.1, x$target, sum, na.rm=T))
pooled_DEGs <- lapply(split(df_DE_ct,df_DE_ct$target), function(x) unique(unlist(lapply(split(x,x$celltype), function(x) x$gene[which(x$padj<0.1)]))))
summary_DEGs_ct <- data.frame(num = lengths(pooled_DEGs),
                              cor = sapply(names(pooled_DEGs), function(x) cor(df_DE_ct$coef[df_DE_ct$target==x & df_DE_ct$celltype=="mural" & df_DE_ct$gene %in% pooled_DEGs[[x]]],
                                                                               df_DE_ct$coef[df_DE_ct$target==x & df_DE_ct$celltype=="endothelial" & df_DE_ct$gene %in% pooled_DEGs[[x]]], use="complete.obs")),
                              p = sapply(names(pooled_DEGs), function(x){
                                test = try(cor.test(df_DE_ct$coef[df_DE_ct$target==x & df_DE_ct$celltype=="mural" & df_DE_ct$gene %in% pooled_DEGs[[x]]], df_DE_ct$coef[df_DE_ct$target==x & df_DE_ct$celltype=="endothelial" & df_DE_ct$gene %in% pooled_DEGs[[x]]], use="complete.obs"))
                                return(ifelse(sum(class(test)=="try-error")>0, 1, test$p.value))
                              }))


# EC
DE_ct_prob_MECOM_EC <- df_DE_ct %>% filter(celltype == "endothelial", target == "MECOM", padj < 0.1)
rownames(DE_ct_prob_MECOM_EC) <- DE_ct_prob_MECOM_EC$gene
DE_ct_prob_MECOM_EC1 <- DE_ct_prob_MECOM_EC %>% select(coef)

rownames(DE_wilcoxauc_cl8) <- DE_wilcoxauc_cl8$feature
DE_wilcoxauc_cl81 <- DE_wilcoxauc_cl8 %>% filter(padj < 0.05) %>% select(logFC)

genes2cor_EC <- intersect(rownames(DE_wilcoxauc_cl81), rownames(DE_ct_prob_MECOM_EC1))
DE_ct_prob_MECOM_EC1 <- DE_ct_prob_MECOM_EC1 %>% filter(rownames(DE_ct_prob_MECOM_EC1) %in% genes2cor_EC)
DE_wilcoxauc_cl81 <- DE_wilcoxauc_cl81 %>% filter(rownames(DE_wilcoxauc_cl81) %in% genes2cor_EC)

df <- DE_ct_prob_MECOM_EC1 %>% mutate(DE_wilcoxauc_cl81)
# Check the distribution
hist(df$logFC, breaks = 100)
hist(df$coef, breaks = 100)

# Calcualte the correlation with Spearman
corr2ref_cl <- cor(DE_ct_prob_MECOM_EC1, DE_wilcoxauc_cl81, method="pearson")
# 0.7804679

# Plot as a scatterplot
df$feature <- rownames(df) 

genes <- c("IGFBP5","NRP2","CRHBP","CAVIN2","TIMP3","GJA5","TAGLN2","EMP1","SQLE")
genes <- df %>% filter(feature %in% genes)

pdf("Plots/MECOM_DE_CROPseq_coef_vs_KO_logFC_endothelial_cells.pdf", height = 4, width = 4)
ggplot(df, aes(x = logFC, y = coef, label = feature)) +
  geom_point(colour = "black", fill= "black", alpha = 0.5, size = 1) +
  geom_point(data = genes, size = 1.5,shape = 21,fill = "#B03A2E", colour = "#B03A2E") +
  theme_classic() + 
  geom_vline(xintercept = 0,  linetype="dotted") +
  geom_hline(yintercept = 0,  linetype="dotted") +
  geom_text_repel(data = genes, mapping = aes(logFC, coef, label = feature), size = 3, max.overlaps = 50) +
  xlab(expression("log"[2]*"FC of DEGs in MECOM KO vs WT in endothelial cells")) + 
  ylab(expression("ANOVA coefficient of DEGs in MECOM gRNA cells vs CTRL in cluster 8"))
dev.off()



# Mural
DE_ct_prob_MECOM_mural <- df_DE_ct %>% filter(celltype == "mural", target == "MECOM", padj < 0.1)
rownames(DE_ct_prob_MECOM_mural) <- DE_ct_prob_MECOM_mural$gene
DE_ct_prob_MECOM_mural1 <- DE_ct_prob_MECOM_mural %>% select(coef)

rownames(DE_wilcoxauc_cl2) <- DE_wilcoxauc_cl2$feature
DE_wilcoxauc_cl21 <- DE_wilcoxauc_cl2 %>% filter(padj < 0.05) %>% select(logFC)

genes2cor_mural <- intersect(rownames(DE_wilcoxauc_cl21), rownames(DE_ct_prob_MECOM_mural1))
DE_ct_prob_MECOM_mural1 <- DE_ct_prob_MECOM_mural1 %>% filter(rownames(DE_ct_prob_MECOM_mural1) %in% genes2cor_mural)
DE_wilcoxauc_cl21 <- DE_wilcoxauc_cl21 %>% filter(rownames(DE_wilcoxauc_cl21) %in% genes2cor_mural)


df <- DE_ct_prob_MECOM_mural1 %>% mutate(DE_wilcoxauc_cl21)
# Check the distribution
hist(df$logFC, breaks = 100)
hist(df$coef, breaks = 100)

# Calcualte the correlation with Pearson
corr2ref_cl <- cor(DE_ct_prob_MECOM_mural1, DE_wilcoxauc_cl21, method="pearson")
# 0.9098838

# Plot as a scatterplot
df$feature <- rownames(df) 

genes <- c("MECOM","PBX3","GOLM1","MAN1A1", "PTN")
genes <- df %>% filter(feature %in% genes)

pdf("Plots/MECOM_DE_CROPseq_coef_vs_KO_logFC_mural.pdf", height = 4, width = 4)
ggplot(df, aes(x = logFC, y = coef, label = feature)) +
  geom_point(colour = "black", fill= "black", alpha = 0.5, size = 1) +
  geom_point(data = genes, size = 1.5,shape = 21,fill = "#B03A2E", colour = "#B03A2E") +
  theme_classic() + 
  geom_vline(xintercept = 0,  linetype="dotted") +
  geom_hline(yintercept = 0,  linetype="dotted") +
  geom_text_repel(data = genes, mapping = aes(logFC, coef, label = feature), size = 3, max.overlaps = 50) +
  xlab(expression("log"[2]*"FC of DEGs in MECOM KO vs WT in mural cells")) + 
  ylab(expression("ANOVA coefficient of DEGs in MECOM gRNA cells vs CTRL in cluster 2"))
dev.off()

```

# Arteriovenous score of ECs

```{r compare to primary endothelial cells in guttube}

seurat_fetal_endo <- readRDS("/links/groups/treutlein/USERS/zhisong_he/Work/intestine_organoids/fetal_atlas_endothelial.seurat.rds")
seurat_fetal_endo <- subset(seurat_fetal_endo, cells = colnames(seurat_fetal_endo)[which(!seurat_fetal_endo$celltype %in% c("COL1A2+","Lymphatic EC"))]) %>% FindVariableFeatures(nfeatures = 3000)
ref_fetal_endo_ct <- sapply(levels(droplevels(seurat_fetal_endo$celltype)), function(ct) rowMeans(seurat_fetal_endo@assays$RNA@data[setdiff(VariableFeatures(seurat_fetal_endo),blacklist),which(seurat_fetal_endo$celltype == ct)]))
avg_expr_fetal_endo <- sapply(levels(droplevels(seurat_fetal_endo$celltype)), function(ct) rowMeans(seurat_fetal_endo@assays$RNA@data[,which(seurat_fetal_endo$celltype == ct)]))

```

```{r A-V score}

DE_fetal_av <- wilcoxauc(seurat_fetal_endo, "celltype", groups_use=c("Arterial EC","Venous EC"))
DEG_fetal_av <- lapply(split(DE_fetal_av, DE_fetal_av$group), function(x){
  x <- x[which(x$padj < 0.01 & x$logFC > log(1.2) & x$auc > 0.7 & x$pct_in > 20 & x$pct_out < 20 & x$pct_in - x$pct_out > 20),]
  x <- x[order(x$auc,decreasing=T),]
  x <- x[which(!x$feature %in% blacklist),]
  return(x)
})
ec <- AddModuleScore(ec, features = list(DEG_fetal_av[['Arterial EC']]$feature), name="arterial")
ec <- AddModuleScore(ec, features = list(DEG_fetal_av[['Venous EC']]$feature), name="venous")
ec$arterial_venous <- ec$arterial1 - ec$venous1
FeaturePlot(ec, c("arterial1","venous1","arterial_venous"), reduction="umap_css", cols=pinknogray_colscheme(30)) & NoAxes() & NoLegend()
png("Plots/plot.umap_css_av_specificity_art_minus_ven.png", height=6, width=8, unit="cm", res=500)
par(mar=c(1,1,1,1), cex=0.5)
plotFeature(Embeddings(ec, "umap_css"), sign(ec$arterial_venous)*sqrt(abs(ec$arterial_venous)), cex=0.9, colorPal=bluered_colscheme, pt_border=T, lwd_border=0.1)
dev.off()
png("Plots/plot.umap_css_av_scores.png", height=6, width=16, unit="cm", res=500)
layout(matrix(1:2,nrow=1)); par(mar=c(1,1,1,1), cex=0.5)
plotFeature(Embeddings(ec, "umap_css"), ec$arterial1, cex=0.9, colorPal=rev(purplegreen_colscheme(30)), pt_border=T, lwd_border=0.1)
plotFeature(Embeddings(ec, "umap_css"), ec$venous1, cex=0.9, colorPal=rev(purplegreen_colscheme(30)), pt_border=T, lwd_border=0.1)
dev.off()

```

# Pathway analysis PORGEN y on all cells
```{r pathway analysis PROGENy}

library(decoupleR)
library(tibble)
library(tidyr)
library(pheatmap)
library(rcartocolor)

net <- get_progeny(organism = 'human', top = 100)
net

# Extract the normalized log-transformed counts
mat_seurat <- as.matrix(seurat@assays$RNA@data)

# Run wmean
acts <- run_wmean(mat=mat_seurat, net=net, .source='source', .target='target',
                  .mor='weight', times = 100, minsize = 5)
acts

# Extract norm_wmean and store it in pathwayswmean in data
seurat[['pathwayswmean']] <- acts %>%
  filter(statistic == 'norm_wmean') %>%
  pivot_wider(id_cols = 'source', names_from = 'condition',
              values_from = 'score') %>%
  column_to_rownames('source') %>%
  Seurat::CreateAssayObject(.)

# Change assay
DefaultAssay(object = seurat) <- "pathwayswmean"
seurat$cluster_condition <- paste0(seurat$RNA_css_snn_res.0.4, "_", seurat$condition)
Idents(seurat) <- "cluster_condition"

# Scale the data
seurat <- ScaleData(seurat)
seurat@assays$pathwayswmean@data <- seurat@assays$pathwayswmean@scale.data

p1 <- DimPlot(seurat, reduction = "umap", label = TRUE, pt.size = 0.5) + 
  NoLegend() + ggtitle('Cell types')
p2 <- (FeaturePlot(seurat, features = c("Trail")) & 
  scale_colour_gradient2(low = 'blue', mid = 'white', high = 'red')) +
  ggtitle('Trail activity')
p1 | p2

# Extract activities from object as a long dataframe
df_seurat <- t(as.matrix(seurat@assays$pathwayswmean@data)) %>%
  as.data.frame() %>%
  mutate(cluster = Idents(seurat)) %>%
  pivot_longer(cols = -cluster, names_to = "source", values_to = "score") %>%
  group_by(cluster, source) %>%
  summarise(mean = mean(score))

# Transform to wide matrix
top_acts_mat_seurat <- df_seurat %>%
  pivot_wider(id_cols = 'cluster', names_from = 'source',
              values_from = 'mean') %>%
  column_to_rownames('cluster') %>%
  as.matrix()

# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("#710085", "#E0E2DD","#006b02"))(palette_length)

my_breaks <- c(seq(-2, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 2, length.out=floor(palette_length/2)))

# Plot
pdf("/home/marinani/PhD_Projects/Vascular_Organoids/Analysis/MECOM_KO/MECOM_KO_day7_day21_together/Plots/pathway_score_progeny.pdf")
pheatmap(top_acts_mat_seurat, border_color = NA, color=my_color, breaks = my_breaks, scale = "column")
dev.off()

```
