---
title: "Brain BVOs HTO assignment"
author: "Marina Nikolova"
date: "11/08/2022"
output: html_document
---

# Basic setup
## Load packages
```{r load packages}

library(Seurat)
library(tidyverse)
library(dplyr)

source("~/feature_plots.r")

```

## Read in the data
```{r reading the data}

# Load the mRNA count matrices
rep1.umi <- Read10X("~/rep1/outs/filtered_feature_bc_matrix/")
rep2.umi <- Read10X("!/rep2/outs/filtered_feature_bc_matrix/")

# Load the HTO count matrices
rep1.hto <- readRDS("~/rep1_hto_mtx.rds")
rep2.hto <- readRDS("~/rep2_hto_mtx.rds")

# Check the number of cells
ncol(rep1.umi)
ncol(rep2.umi)

```

```{r prepare mRNA matrices}

# The column names (cells) have an additional "-1" which is not present in the HTO counts
# To allow for joining of the common barcodes in both matrices, we rename the barcodes by removing the additional "-1" pattern
colnames_rep1 <- colnames(rep1.umi)
colnames_rep1 <- gsub("-1", "", colnames_rep1)
colnames(rep1.umi) <- colnames_rep1

colnames_rep2 <- colnames(rep2.umi)
colnames_rep2 <- gsub("-1", "", colnames_rep2)
colnames(rep2.umi) <- colnames_rep2

```


## Generate joint barcode matrices
```{r joint matrices}

# Select cell barcodes detected by both RNA and HTO 
rep1.joint.bcs <- intersect(colnames(rep1.umi), colnames(rep1.hto))
rep2.joint.bcs <- intersect(colnames(rep2.umi), colnames(rep2.hto))

# Check how many cells have an assigned barcode
table(colnames(rep1.umi) %in% colnames(rep1.hto))
table(colnames(rep2.umi) %in% colnames(rep2.hto))

# Subset RNA and HTO counts by joint cell barcodes
rep1.umi <- rep1.umi[, rep1.joint.bcs]
rep2.umi <- rep2.umi[, rep2.joint.bcs]

rep1.hto <- as.matrix(rep1.hto[, rep1.joint.bcs])
rep2.hto <- as.matrix(rep2.hto[, rep2.joint.bcs])

# Confirm that the HTO have the correct names
rownames(rep1.hto)
rownames(rep2.hto)

```

## Setup Seurat object and add in the HTO data
```{r setup seurat with HTO}

# Setup Seurat object
rep1.hashtag <- CreateSeuratObject(counts = rep1.umi, project = "Replicate1")
rep2.hashtag <- CreateSeuratObject(counts = rep2.umi, project = "Replicate2")

# Normalize, find variable features, scale the data and run PCA on the RNA data with log normalization
rep1.hashtag <- NormalizeData(rep1.hashtag) 
rep2.hashtag <- NormalizeData(rep2.hashtag) 

```

# Demultiplex cells based on HTO enrichment
## Add HTO as an indepent assay to the Seurat object
```{r add HTO data}

# Add HTO data as a new assay independent from RNA
rep1.hashtag[["HTO"]] <- CreateAssayObject(counts = rep1.hto)
rep2.hashtag[["HTO"]] <- CreateAssayObject(counts = rep2.hto)

# Normalize HTO data, here we use centered log-ratio (CLR) transformation
rep1.hashtag <- NormalizeData(rep1.hashtag, assay = "HTO", normalization.method = "CLR", margin = 2)
rep2.hashtag <- NormalizeData(rep2.hashtag, assay = "HTO", normalization.method = "CLR", margin = 2)

```

## Demultiplex using the Seurat function HTODemux() to assign single cells back to their sample origins
```{r demultiplex}

# If one has a very large dataset Seurat suggests using k_function = 'clara'. 
# This is a k-medoid clustering function for large applications 
# One can also play with additional parameters (see documentation for HTODemux()) to adjust the threshold for classification 
# Here I am using the default settings
rep1.hashtag <- HTODemux(rep1.hashtag, assay = "HTO", positive.quantile = 0.99)
rep2.hashtag <- HTODemux(rep2.hashtag, assay = "HTO", positive.quantile = 0.99)

```

# Visualization
## Output from running HTODemux() is saved in the object metadata
## We can visualize how many cells are classified as singlets, doublets and negative/ambiguous cells
```{r visualize HTO classification}

# Global classification results
table(rep1.hashtag$HTO_classification.global)
table(rep2.hashtag$HTO_classification.global)

# Change the order to display the hashtags 
hto_order <- c("Doublet", "HTO1", "HTO2", "HTO3", "HTO4", "HTO5", "HTO6", "HTO7", "HTO8", "HTO9", "HTO10", "HTO12", "HTO13", "HTO14", "Negative")
rep1.hashtag@meta.data$hash.ID <- factor(rep1.hashtag@meta.data$hash.ID, levels = hto_order)
rep2.hashtag@meta.data$hash.ID <- factor(rep2.hashtag@meta.data$hash.ID, levels = hto_order)

# Group cells based on the max HTO signal
rep1.hashtag@meta.data$HTO_maxID <- factor(rep1.hashtag@meta.data$HTO_maxID, levels = hto_order)
Idents(rep1.hashtag) <- "HTO_maxID"
RidgePlot(rep1.hashtag, assay = "HTO", features = rownames(rep1.hashtag[["HTO"]])[1:12], ncol = 4)

rep2.hashtag@meta.data$HTO_maxID <- factor(rep2.hashtag@meta.data$HTO_maxID, levels = hto_order)
Idents(rep2.hashtag) <- "HTO_maxID"
RidgePlot(rep2.hashtag, assay = "HTO", features = rownames(rep2.hashtag[["HTO"]])[1:12], ncol = 4)

```
## Visualize pairs of HTO signals to confirm mutual exclusivity in singlets

```{r scatter plot}

# Check a few random 
FeatureScatter(rep1.hashtag, feature1 = "hto_HTO1", feature2 = "hto_HTO2")
FeatureScatter(rep1.hashtag, feature1 = "hto_HTO5", feature2 = "hto_HTO7")
FeatureScatter(rep2.hashtag, feature1 = "hto_HTO1", feature2 = "hto_HTO2")
FeatureScatter(rep2.hashtag, feature1 = "hto_HTO5", feature2 = "hto_HTO7")

```

## Compare number of UMIs for singlets, doublets and negative cells
```{r QC of HTOs}

global.order <- c("Singlet", "Doublet", "Negative")
rep1.hashtag@meta.data$HTO_classification.global <- factor(rep1.hashtag@meta.data$HTO_classification.global, levels = global.order)
rep2.hashtag@meta.data$HTO_classification.global <- factor(rep2.hashtag@meta.data$HTO_classification.global, levels = global.order)

# Assign the droplets singlets, doublets or negative identities
Idents(rep1.hashtag) <- "HTO_classification.global"
Idents(rep2.hashtag) <- "HTO_classification.global"

# Plot UMI counts 
VlnPlot(rep1.hashtag, features = "nCount_RNA", pt.size = 0.1, log = TRUE)
VlnPlot(rep2.hashtag, features = "nCount_RNA", pt.size = 0.1, log = TRUE)

```

## Generate a two dimensional tSNE embedding for HTOs.Here we are grouping cells by singlets and doublets for simplicity
```{r 2D embedding of HTOs}

# First, we will remove negative cells from the object
rep1.hashtag.subset <- subset(rep1.hashtag, idents = "Negative", invert = TRUE)
rep2.hashtag.subset <- subset(rep2.hashtag, idents = "Negative", invert = TRUE)

# Calculate a tSNE embedding of the HTO data
# For each replicate separately
DefaultAssay(rep1.hashtag.subset) <- "HTO"
DefaultAssay(rep2.hashtag.subset) <- "HTO"

rep1.hashtag.subset <- ScaleData(rep1.hashtag.subset, features = rownames(rep1.hashtag.subset), verbose = FALSE) %>% 
                      RunPCA(features = rownames(rep1.hashtag.subset), approx = FALSE) %>%
                      RunTSNE(dims = 1:8, perplexity = 100)
rep2.hashtag.subset <- ScaleData(rep2.hashtag.subset, features = rownames(rep2.hashtag.subset), verbose = FALSE) %>% 
                      RunPCA(features = rownames(rep2.hashtag.subset), approx = FALSE) %>%
                      RunTSNE(dims = 1:8, perplexity = 100)

Idents(rep1.hashtag.subset) <- "HTO_classification.global"
Idents(rep2.hashtag.subset) <- "HTO_classification.global"
DimPlot(rep1.hashtag.subset) | DimPlot(rep2.hashtag.subset) 

# One can also visualize the more detailed classification result by running Idents(object) <- 'HTO_classification' before plotting. 
# Here, one can see that each of the small clouds on the tSNE plot corresponds to one of the 28 possible doublet combinations.
Idents(rep1.hashtag.subset) <- 'hash.ID' 
Idents(rep2.hashtag.subset) <- 'hash.ID' 
( DimPlot(rep1.hashtag.subset) | DimPlot(rep2.hashtag.subset) )

```
## Create an HTO heatmap
```{r HTO heatmap}

# To increase the efficiency of plotting, you can subsample cells using the num.cells argument
HTOHeatmap(rep1.hashtag, assay = "HTO", ncells = 5000)
HTOHeatmap(rep2.hashtag, assay = "HTO", ncells = 5000)

```

## Prepare the mRNA transcript data for analysis and visualization
```{r}

# Extract the singlets
rep1.singlet <- subset(rep1.hashtag, idents = "Singlet")
rep2.singlet <- subset(rep2.hashtag, idents = "Singlet")

```
